---
title: "Deep-dive · The ABCD-Shaped Dataset"
subtitle: "The dataset used in Modules 2, 3 and 4"
pagetitle: "ABCD-Shaped Dataset & Data Dictionary | ABCD R Workshop"
description-meta: "A synthetic longitudinal dataset shaped like an ABCD release: 500 participants across 5 sessions, 53 variables, a searchable data dictionary and coded factor levels."
open-graph:
  description: "A synthetic ABCD-shaped dataset with a searchable data dictionary — 500 participants, 5 sessions, 53 variables, structured missing data."
twitter-card:
  description: "A synthetic ABCD-shaped dataset with a searchable data dictionary — 500 participants, 5 sessions, 53 variables, structured missing data."
---

```{r}
#| label: page-setup
#| include: false
library(reactable)
source("R/reactable_utils.R")
```

Modules 2, 3 and 4 all run on one file: `data/data.RDS`. It is bigger and messier
than the Module 1 dataset, and deliberately so — it is built to behave like a real
ABCD release, down to the variable names, the coded factor levels, and the shape of
the missing data.

This page is the reference for it. Come back here whenever you need to look up what
a column means.

::: {.callout-important}
## The data is not real
Every row is simulated. No real participant contributes a row, a value or a record.
The *shape* was fitted from a de-identified ABCD extract — column names, variable
types, level sets, means, SDs and missingness rates — and the extract was then set
aside. Only those summary numbers live in the generator,
[`R/generate-abcd-shaped-data.R`](https://github.com/beepboopstats/r-stats-workshop/blob/main/R/generate-abcd-shaped-data.R).
:::

## Get the files

Three files. Put all three in the `data/` folder of your RStudio Project — the same
place `abcd-synthetic.csv` went in [Module 0](module0.qmd#set-up-a-project-folder).

[data.RDS](../data/data.RDS){.btn .btn-outline-primary .btn-sm}
[data_dictionary.csv](../data/data_dictionary.csv){.btn .btn-outline-primary .btn-sm}
[data_dictionary_levels.csv](../data/data_dictionary_levels.csv){.btn .btn-outline-primary .btn-sm}

| File | What it holds |
|:--|:--|
| `data.RDS` | The data itself — one row per participant per visit |
| `data_dictionary.csv` | One row per **variable**: what it is, who reported it, which instrument it came from |
| `data_dictionary_levels.csv` | One row per **level** of every coded variable: `1` means Male, `2` means Female, and so on |

A real ABCD release arrives exactly like this: data in one place, meaning in
another. Learning to work with the pair is part of the point.

## Load it

```{r}
#| label: load
library(readr)
library(dplyr)

data <- readRDS("data/data.RDS")

dict <- read_csv("data/data_dictionary.csv")
levels_dict <- read_csv("data/data_dictionary_levels.csv")
```

`.RDS` is R's own file format, so `readRDS()` needs no package and takes no
arguments beyond the path. It also preserves everything: factors stay factors, and
the visit timestamp stays a date-time. A CSV would lose both, and you would have to
rebuild them by hand every time you opened the file.

```{r}
#| label: dims
dim(data)
```

**2,295 rows and 53 columns** — 500 participants, each with up to five visits.

## The shape

One row is **one participant at one visit**. Two columns identify that row:

```{r}
#| label: keys
data |>
  select(participant_id, session_id, ab_g_dyn__visit_age) |>
  head(5)
```

Neither column identifies a row on its own — you need the pair. That is what makes
this a **longitudinal** dataset, and it is the reason nearly everything in Modules 3
and 4 has to think about the fact that the five rows belonging to `sub-0001` are not
five independent observations.

The five sessions are annual visits:

| `session_id` | Visit |
|:--|:--|
| `ses-00A` | Baseline |
| `ses-01A` | 1-year follow-up |
| `ses-02A` | 2-year follow-up |
| `ses-03A` | 3-year follow-up |
| `ses-04A` | 4-year follow-up |

```{r}
#| label: sessions
count(data, session_id)
```

The counts fall because participants drop out — 500 at baseline, 414 still present
four years later. Nobody who leaves comes back.

## How ABCD names its variables

The column names look forbidding until you know they are structured. Read them
left to right, splitting on the **double** underscore:

```
nt_y_stq__screen__wkdy_sum
│  │ │    │       │
│  │ │    │       └── the statistic: weekday sum
│  │ │    └────────── the scale: screen time
│  │ └─────────────── the instrument: Screen Time Questionnaire
│  └───────────────── who reported it: y = youth, p = parent, g = general
└──────────────────── the domain: nt = Novel Technologies
```

Eight domain prefixes appear in this dataset:

| Prefix | Domain |
|:--|:--|
| `ab` | ABCD (General) — IDs, site, visit and demographic variables |
| `fc` | Friends, Family, & Community |
| `mh` | Mental Health |
| `mr` | Imaging (structural MRI) |
| `nc` | Neurocognition |
| `nt` | Novel Technologies (screen use) |
| `ph` | Physical Health |
| `su` | Substance Use |

Once that clicks, `mh_p_cbcl__synd__int_sum` reads as *Mental Health · parent-reported ·
Child Behavior Checklist · syndrome scale · internalizing · sum* without looking
anything up. It also means `select()` helpers do real work — `starts_with("nc_")`
gives you every cognition variable at once.

## Browse the variables

All 53 columns, grouped by domain. Search, filter, sort, expand a domain, or
download the whole table as a CSV.

```{r}
#| label: dict-table
#| echo: false
tibble(name = names(data)) |>
  left_join(select(dict, name, domain, source, label), by = "name") |>
  mutate(
    domain = if_else(is.na(domain), "Identifiers", domain),
    source = if_else(is.na(source), "General", source),
    label  = case_when(
      name == "participant_id" ~ "Participant identifier",
      name == "session_id"     ~ "Session (visit) identifier",
      .default = label
    )
  ) |>
  grouped_reactable(
    group_var = "domain",
    id = "data-dict",
    columns = list(
      name   = reactable::colDef(maxWidth = 260),
      domain = reactable::colDef(maxWidth = 200),
      source = reactable::colDef(maxWidth = 110),
      label  = reactable::colDef(minWidth = 300)
    )
  )
```

`source` is **who answered**: `Youth` is the child's own report, `Parent` the
caregiver's, `General` an administrative field. Parent and youth reports of the
*same* construct disagree — this dataset has both a parent and a youth screen-time
measure, and comparing them is a Module 3 exercise waiting to happen.

The fifteen instruments behind those variables:

```{r}
#| label: instruments
#| echo: false
dict |>
  distinct(table_name, table_label, source, domain) |>
  arrange(domain, table_name) |>
  select(instrument = table_name, `what it is` = table_label, source, domain) |>
  knitr::kable()
```

## Coded values, and their labels

Here is the thing that surprises everyone the first time. Look at sex:

```{r}
#| label: coded-sex
count(data, ab_g_stc__cohort_sex)
```

The values are `1` and `2`. Not `"Male"` and `"Female"` — the meaning is not in the
data file. It is in `data_dictionary_levels.csv`, which is why that file matters:

```{r}
#| label: sex-levels
levels_dict |>
  filter(name == "ab_g_stc__cohort_sex")
```

This is not sloppiness. Real ABCD releases ship coded values and a separate
dictionary on purpose: codes survive translation, they do not break when a label is
reworded, and the same file can serve English and Spanish labels (hence the
`label_es` column). The cost is that **you must apply the labels yourself**.

### Labelling one variable

Pull that variable's levels out of the dictionary, then hand them to `factor()`:

```{r}
#| label: label-sex
sex_levels <- levels_dict |>
  filter(name == "ab_g_stc__cohort_sex") |>
  arrange(order_level)

data <- data |>
  mutate(
    ab_g_stc__cohort_sex = factor(
      ab_g_stc__cohort_sex,
      levels = sex_levels$value,   # the codes stored in the data
      labels = sex_levels$label    # what each code means
    )
  )

count(data, ab_g_stc__cohort_sex)
```

Three things worth noticing:

- **`levels` and `labels` must line up**, position by position. `arrange(order_level)`
  is what guarantees that — the dictionary already knows the order these levels
  should appear in, which matters for ordinal variables like income bands, where
  alphabetical order would put `> 100k` before `< 50k`.
- The variable is still a **factor** afterwards, not text. It plots and tabulates in
  the dictionary's order rather than alphabetically.
- Any code present in `levels` but absent from the data simply produces a zero
  count. That is usually what you want — `Decline to answer` is a real category even
  in a wave where nobody chose it.

Repeat that block for each variable you actually use. Doing all thirteen coded
variables in one stroke is a Module 2 topic — it needs tools this page has not
introduced yet, and typing it out longhand a few times first is not wasted effort.

### Every coded variable

The thirteen variables that carry codes, and what each code means. Expand a
variable to see its levels.

```{r}
#| label: levels-table
#| echo: false
levels_dict |>
  arrange(name, order_level) |>
  select(name, value, label) |>
  grouped_reactable(
    group_var = "name",
    id = "levels-dict",
    columns = list(
      name  = reactable::colDef(maxWidth = 320),
      value = reactable::colDef(maxWidth = 110),
      label = reactable::colDef(minWidth = 300)
    )
  )
```

Two codes recur across variables and are **not** ordinary responses:

| Code | Meaning |
|:--|:--|
| `777` | Decline to answer |
| `999` | Don't know |

They are stored as ordinary levels, so a mean or an ordinal model that treats them
as the top of the scale will be badly wrong. Deciding what to do with them — drop
them, or keep them as their own category — is an analysis choice you have to make
explicitly. Module 3 makes it.

## What is missing, and why

About **27% of all cells are empty**, and that is realistic rather than broken. What
matters is that the missingness is *structured*: it is never a uniform random
sprinkle, so dropping incomplete rows is never free.

```{r}
#| label: missing-by-column
colMeans(is.na(data)) |>
  sort(decreasing = TRUE) |>
  head(8) |>
  round(2)
```

Four different mechanisms produce those gaps:

- **Instrument schedules.** Several measures are only collected at baseline, the
  2-year and the 4-year visit — so they are ~40% missing by design, not by accident.
  A few NIH Toolbox subtests were retired after baseline; the Family Environment
  cohesion scale was only added at the 4-year visit.
- **Visit type.** Anthropometrics are ~80% missing at `Remote` visits and ~6%
  missing elsewhere. Nobody can measure a waist circumference over video.
- **Data quality.** MRI failure rises with age, because older children move more in
  the scanner.
- **Attrition.** 86 participants never reach the 4-year visit — and they are not a
  random 86. They had higher baseline externalizing scores and were more often in
  the lowest income band.

That last one is the important one. It means a complete-case analysis quietly
changes *who* you are studying, and that is a point Module 3 has to make with this
dataset rather than in the abstract.

## Is there anything in here to find?

Yes. The dataset is not noise — real associations were built into it, sized so that
a straightforward analysis at n = 500 will actually recover them. Finding them is
the work of Modules 3 and 4.

::: {.callout-warning collapse="true"}
## Spoilers — the full list of planted effects

Open this if you are teaching from the dataset, or working through it alone and
want to check your answer. If you are attending Modules 3 and 4, you will get more
out of them by leaving it shut.

- **E1** Age → youth-reported screen time. About +1.0 h/day per year on weekdays,
  +1.8 h/day per year at weekends.
- **E2** Screen time → CBCL internalizing, holding age and sex constant. About
  +0.27 points per weekday hour.
- **E3** Household income → NIH Toolbox crystallized, picture vocabulary and oral
  reading. About 0.6 SD between the lowest and the highest income group at baseline.
- **E4** Sex. Males about +1.5 externalizing, females about +1.1 internalizing,
  males about 7% larger total cortical volume.
- **E5** Site. Random intercepts on cognition (SD 2.5 standard-score points) and on
  screen time (SD 0.4 h) — the reason site belongs in a model rather than being
  ignored.
- **E6** Family conflict (FES) → CBCL externalizing, about +2.9 points across the
  range of the conflict mean.
- **E7** Parental monitoring → less substance-use curiosity; age → more curiosity.
- **E8** Age → cortical thinning, about −0.030 mm/year, and a slow decline in total
  cortical volume after age 12.
- **E9** Household income → perceived neighbourhood safety. 53% of the highest
  income group strongly agree their neighbourhood is safe, against 30% of the
  lowest.
- **E10** Missingness is structured, as described above — instrument schedules,
  remote visits, age-related MRI failure, and MNAR attrition.

Numbers are what a straightforward model recovers from the file as it stands, not
the coefficients fed to the simulation.
:::

## Where this is used

- [Module 2](module2.qmd) — exploring and visualising it
- [Module 3](module3.qmd) — testing hypotheses on it
- [Module 4](module4.qmd) — running an analysis end to end on it

[Module 1](module1.qmd) does **not** use this file. It uses the much smaller
`data/abcd-synthetic.csv`, which has plain English column names and three visits —
deliberately, so that a first hour in R is spent learning R rather than decoding
variable names.

## Real ABCD data

Everything on this page mirrors the real thing, but the real thing needs
authorisation. Access runs through the
[NIMH Data Archive](https://nda.nih.gov/abcd), and the official variable
documentation lives at
[data.abcdstudy.org](https://data.abcdstudy.org/reports/abcd-docs-next/documentation/non_imaging/index.html) —
the `url_table` column of `data_dictionary.csv` points at the relevant page for
every instrument here.
