---
title: "Deep-dive · Writing Your Own Functions"
subtitle: "Optional companion to Module 2"
pagetitle: "Writing R Functions for dplyr & ggplot2 | ABCD R Workshop"
description-meta: "Turn copy-pasted dplyr pipelines and ggplot2 code into reusable R functions: arguments, return values, {{ }} embracing, plot functions and layer bundles."
open-graph:
  description: "Refactor Module 2's dplyr and ggplot2 code into reusable R functions, step by step."
twitter-card:
  description: "Refactor Module 2's dplyr and ggplot2 code into reusable R functions, step by step."
editor_options:
  chunk_output_type: console
---

[Module 2](module2.qmd) built every figure by copying the one before and changing
a few lines. That is the right way to *discover* an analysis. It is the wrong way
to *keep* one. This page takes the Module 2 code and turns it, one step at a
time, into a handful of small functions — until the whole session fits in about
six lines.

```{r}
#| label: setup
library(dplyr)
library(ggplot2)

data <- readRDS("data/data.RDS")
levels_dict <- readr::read_csv("data/data_dictionary_levels.csv")
```

## Why bother

Look back at Module 2's Exercise 1. Its solution is the Q1 facet pipeline with
three lines changed. Exercise 2 is the Q2 pipeline with one measure swapped. Two
copies, each drifting slightly away from the original — and one real bug that
drifted in: Q1 sets the order of `sex` and Q2 does not, so **the colours swap
between the two figures**.

Copies go wrong in exactly this way. Fix a problem in one and you have to
remember every other place it was pasted. A **function** is a named piece of
code you write once and call many times. Fix it once, and every call is fixed.

A common rule of thumb: once you have copied a block of code twice, make it a
function.

## The anatomy of a function

```{r}
#| label: first-function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}
```

Four parts:

- **`se`** — the name, assigned with `<-` like any other object
- **`function(x)`** — the **arguments**: the inputs, and the names they go by
  *inside* the function
- **`{ ... }`** — the **body**: the code that runs
- the **return value** — whatever the last line of the body produces. No
  `return()` needed

That function is the standard error from Module 2 — SD divided by the square root
of n — written so it copes with missing values: `sum(!is.na(x))` counts only the
values that are present. Call it like any built-in function:

```{r}
#| label: use-se
se(c(4, 8, 6, NA, 5))

sub_screen <- data |>
  group_by(session_id) |>
  summarise(
    mean = mean(nt_y_stq__screen__wkdy_sum, na.rm = TRUE),
    se   = se(nt_y_stq__screen__wkdy_sum)
  )
sub_screen
```

### Default arguments

An argument can have a default value, used when the caller does not supply one:

```{r}
#| label: defaults
pct_missing <- function(x, digits = 1) {
  round(100 * mean(is.na(x)), digits)
}

pct_missing(data$ph_y_anthr__waist_001)
pct_missing(data$ph_y_anthr__waist_001, digits = 3)
```

You have been using defaults since Module 1: `mean()` has `na.rm = FALSE`
unless you say otherwise.

::: {.callout-tip}
## You have met one already
The [wrangling deep-dive](module2-wrangling.qmd#labelling-every-coded-variable-at-once)
wrote `apply_labels(x, var)`, which labels any coded variable from the
dictionary. Same anatomy: two arguments, a body, and the last line's value
returned.
:::

## Functions around dplyr: `{{ }}`

`se()` works on a vector. The Module 2 code, though, is pipelines, and the
natural next step — passing a column *name* into a function — fails in a way
that confuses everyone the first time:

```{r}
#| label: naive-dplyr
#| error: true
mean_of <- function(data, var) {
  data |> summarise(mean = mean(var, na.rm = TRUE))
}

mean_of(data, ph_y_anthr__height_mean)
```

Inside a dplyr verb, a bare name like `height` means "the column called
`height`". But inside a function, `var` is just an argument. dplyr looks for a
column literally called `var`, does not find one, and then looks for an ordinary
variable instead — and fails.

The fix is to **embrace** the argument with double braces, `{{ var }}`. It tells
dplyr: "don't use the word `var`; use whatever the caller typed in its place."

```{r}
#| label: embrace
mean_of <- function(data, var) {
  data |> summarise(mean = mean({{ var }}, na.rm = TRUE))
}

mean_of(data, ph_y_anthr__height_mean)
```

That is the one rule to remember: **when an argument is a column name that is
passed into a dplyr or ggplot2 function, wrap it in `{{ }}`.**

### A summary function

Now the Q2 summary — n, mean, SD and SE per group — as a function. The `...`
("dots") argument collects *any number* of extra arguments and passes them on,
here to `group_by()`, so the caller can group by as many columns as they like:

```{r}
#| label: summarise-mean-se
summarise_mean_se <- function(data, var, ...) {
  data |>
    filter(!is.na({{ var }})) |>
    group_by(...) |>
    summarise(
      n    = n(),
      mean = mean({{ var }}),
      sd   = sd({{ var }}),
      se   = sd / sqrt(n),
      .groups = "drop"
    )
}

data |>
  summarise_mean_se(nt_y_stq__screen__wkdy_sum, session_id)
```

One line per summary table, for any variable and any grouping.

## Wrap the data preparation

Module 2's section 3 turned `data` into `sub2`: select and rename, round, recode
sex, build `visit`, join the site labels. Here it is in full as the "before" —
the same code, minus the `glimpse()` calls used to peek during the live
session:

```{r}
#| label: before
site_levels <- levels_dict |>
  filter(
    name == "ab_g_dyn__design_site"
  ) |>
  select(
    value,
    site_label = label
  )

sub2 <- data |>
  select(
    1:2,
    age = ab_g_dyn__visit_age,
    site = ab_g_dyn__design_site,
    income = ab_g_dyn__cohort_income__hhold__3lvl,
    sex = ab_g_stc__cohort_sex,
    ethn = ab_g_stc__cohort_ethn,
    matches("_anthr_"),
    matches("_screen_"),
    matches("_cbcl_")
  ) |>
  rename(
    height = ph_y_anthr__height_mean,
    weight = ph_y_anthr__weight_mean,
    waist = ph_y_anthr__waist_001
  ) |>
  as_tibble() |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x)
    )
  ) |>
  mutate(
    sex = case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female",
      .default = NA
    ),
    visit = stringr::str_replace(session_id, "ses-", "") |>
      readr::parse_number() |>
      factor(),
    site = as.character(site)
  ) |>
  left_join(
    site_levels |>
      mutate(
        value = as.character(value)
      ),
    join_by(site == value)
  )
```

Wrapped in a function, the two pieces — building the site lookup and formatting
the data — travel together, and the inputs become explicit arguments:

```{r}
#| label: format-abcd
format_abcd <- function(data, levels_dict) {
  site_levels <- levels_dict |>
    filter(name == "ab_g_dyn__design_site") |>
    select(value, site_label = label) |>
    mutate(value = as.character(value))

  data |>
    select(
      1:2,
      age    = ab_g_dyn__visit_age,
      site   = ab_g_dyn__design_site,
      income = ab_g_dyn__cohort_income__hhold__3lvl,
      sex    = ab_g_stc__cohort_sex,
      ethn   = ab_g_stc__cohort_ethn,
      matches("_anthr_"),
      matches("_screen_"),
      matches("_cbcl_")
    ) |>
    rename(
      height = ph_y_anthr__height_mean,
      weight = ph_y_anthr__weight_mean,
      waist  = ph_y_anthr__waist_001
    ) |>
    as_tibble() |>
    mutate(across(where(is.numeric), ~ round(.x))) |>
    mutate(
      sex = case_when(
        sex == 1 ~ "Male",
        sex == 2 ~ "Female",
        .default = NA
      ),
      visit = stringr::str_replace(session_id, "ses-", "") |>
        readr::parse_number() |>
        factor(),
      site = as.character(site)
    ) |>
    left_join(site_levels, join_by(site == value))
}
```

`site_levels` is now created *inside* the function. It exists only while the
function runs and cannot clash with anything else in your session — one of the
quiet benefits of functions.

Whenever you refactor, check that nothing changed:

```{r}
#| label: check-identical
identical(format_abcd(data, levels_dict), sub2)
```

`identical()` compares two objects exactly. `TRUE` means the function gives the
same `sub2`, value for value, as the original pipeline.

## Functions that return plots

A ggplot is an object like any other, so a function can build one and return it.
The same `{{ }}` rule applies inside `aes()` and inside tidyselect arguments such
as `select()` and `pivot_longer(cols = )`.

### Reusable layer bundles

First, the pieces every Module 2 figure repeated — the theme, the legend
position, the colours. A function can return a **`list()` of layers**, and
ggplot2 adds every item in the list when you `+` it:

```{r}
#| label: layer-bundles
sex_colours <- c(Male = "orange", Female = "grey60")

scale_sex <- function() {
  list(
    scale_fill_manual(values = sex_colours),
    scale_colour_manual(values = sex_colours)
  )
}

theme_workshop <- function(base_size = 20) {
  list(
    theme_bw(base_size),
    theme(legend.position = "top")
  )
}
```

Note the **named** colour values. `c(Male = "orange", ...)` ties each colour to a
*category* rather than to a position, so males are orange whatever order the
levels are in. That fixes Module 2's colour-swap bug once, for every figure
that uses `scale_sex()`.

### The facet boxplot

```{r}
#| label: plot-boxplots
plot_boxplots <- function(data, cols) {
  data |>
    mutate(sex = factor(sex, levels = c("Male", "Female"))) |>
    select(participant_id, visit, sex, {{ cols }}) |>
    tidyr::pivot_longer(
      cols = {{ cols }},
      names_to = "what",
      values_to = "value"
    ) |>
    mutate(visit = paste0("yr-", visit) |> factor()) |>
    ggplot(aes(x = visit, y = value, fill = sex)) +
    geom_boxplot(col = "black") +
    facet_grid(~ what, scales = "free") +
    coord_flip() +
    scale_sex() +
    theme_workshop()
}
```

Module 2's Q1 figure is now one call:

```{r}
#| label: fig-fn-anthro
#| fig-cap: "Q1 from Module 2, as one function call"
#| fig-alt: "Three panels of horizontal boxplots — height, waist and weight — by visit, orange for males and grey for females."
#| fig-width: 12
#| fig-height: 6
#| warning: false
plot_boxplots(sub2, height:waist)
```

And Exercise 1 is the same call with different columns. `cols` accepts anything
`select()` does — ranges, names, `matches()`:

```{r}
#| label: fig-fn-screen
#| fig-cap: "Module 2's Exercise 1, as one function call"
#| fig-alt: "Four panels of horizontal boxplots of the four screen-time measures by visit, orange for males and grey for females."
#| fig-width: 12
#| fig-height: 6
#| warning: false
plot_boxplots(sub2, matches("_screen_"))
```

### The mean ± SE plot

```{r}
#| label: plot-mean-se
plot_mean_se <- function(summary, x, group, dodge = 0.7) {
  pd <- position_dodge(width = dodge)

  summary |>
    ggplot(aes(x = {{ x }}, y = mean, colour = {{ group }}, fill = {{ group }}, group = {{ group }})) +
    geom_errorbar(aes(ymin = mean - se, ymax = mean + se), position = pd, width = .3) +
    geom_line(position = pd) +
    geom_point(position = pd, size = 7) +
    scale_sex() +
    theme_workshop()
}
```

It expects a table with `mean` and `se` columns — exactly what
`summarise_mean_se()` returns — so the two chain together:

```{r}
#| label: fig-fn-mean-se
#| fig-cap: "Q2 from Module 2, as a short pipeline of function calls"
#| fig-alt: "Mean weekday screen time by age from 10 to 15, one line per sex with error bars, males in orange and females in grey."
#| fig-width: 10
#| fig-height: 6
sub2 |>
  filter(age >= 10 & age <= 15) |>
  mutate(age = factor(age)) |>
  summarise_mean_se(nt_y_stq__screen__wkdy_sum, age, sex) |>
  plot_mean_se(age, sex)
```

Males are orange again, as in Q1 — the named colours at work. The legend still
lists `Female` first, because legend *order* comes from the factor levels, not
from the colour names. Add `mutate(sex = factor(sex, levels = c("Male", "Female")))`
if the order matters.

One small difference from the original: Q2 dropped a row if *any* of four
measures was missing, while `summarise_mean_se()` only needs the one being
summarised. So each group here keeps a few more participants (146 rather than
140 girls at age 10, for instance), and the means shift very slightly. Neither
rule is wrong, but a refactor is a good moment to choose one deliberately.

Because the function returns an ordinary ggplot, you can keep adding layers to
the result:

```{r}
#| label: fig-fn-extend
#| fig-alt: "Mean CBCL externalizing score by age and sex with error bars, with axis titles and a plot title added."
#| fig-width: 10
#| fig-height: 6
sub2 |>
  filter(age >= 10 & age <= 15) |>
  mutate(age = factor(age)) |>
  summarise_mean_se(mh_p_cbcl__synd__ext_sum, age, sex) |>
  plot_mean_se(age, sex) +
  labs(
    x = "Age (years)",
    y = "CBCL externalizing (mean ± SE)",
    title = "Externalizing problems by age and sex"
  )
```

## Putting it together

Everything from Module 2, with the functions defined above:

```r
sub2 <- format_abcd(data, levels_dict)

plot_boxplots(sub2, height:waist)
plot_boxplots(sub2, matches("_screen_"))

sub2 |>
  filter(age >= 10 & age <= 15) |>
  mutate(age = factor(age)) |>
  summarise_mean_se(nt_y_stq__screen__wkdy_sum, age, sex) |>
  plot_mean_se(age, sex)
```

Each line says *what* it does, not *how*. The how lives in one place.

### The same plot for many variables

With column names held as **text**, one more piece of syntax is needed:
`.data[[name]]` means "the column whose name is stored in `name`". It works
anywhere a bare column name would, including through `{{ }}`:

```{r}
#| label: many-plots
outcomes <- c(
  "nt_y_stq__screen__wkdy_sum",
  "mh_p_cbcl__synd__int_sum",
  "mh_p_cbcl__synd__ext_sum"
)

plots <- lapply(outcomes, \(v) {
  sub2 |>
    filter(age >= 10 & age <= 15) |>
    mutate(age = factor(age)) |>
    summarise_mean_se(.data[[v]], age, sex) |>
    plot_mean_se(age, sex) +
    labs(title = v)
})

length(plots)
```

`lapply()` runs the function once for each element of `outcomes` and returns a
list of the results — here, three finished plots. (`\(v)` is shorthand for
`function(v)`.) Look at any one with `plots[[2]]`, or save them all:

```r
for (i in seq_along(plots)) {
  ggsave(
    paste0("output/", outcomes[i], ".png"),
    plot = plots[[i]],
    width = 10, height = 6, dpi = 300
  )
}
```

Adding a fourth outcome is now a one-word change.

## Keeping functions in a file

Functions defined in the middle of an analysis script get lost. Put them in their
own file in your project — for example `R/functions.R` — and load them at the top
of every script that needs them:

```r
library(dplyr)
library(ggplot2)

source("R/functions.R")   # defines format_abcd(), plot_boxplots(), ...
```

`source()` runs every line of a file, so the functions are defined and ready to
use. The same relative-path rule from Module 1 applies: it works because your
RStudio Project sets the working directory.

### Failing helpfully

A function used by other people — or by you in three months — should say clearly
when it is being misused. Without a check, passing the raw data to
`plot_boxplots()` fails deep inside `mutate()` with an error about a column called
`sex`. A check at the top turns that into a message the caller can act on:

```{r}
#| label: fail-helpfully
#| error: true
plot_boxplots <- function(data, cols) {
  if (!"sex" %in% names(data)) {
    stop("`data` needs a `sex` column. Run format_abcd() on it first.")
  }

  data |>
    mutate(sex = factor(sex, levels = c("Male", "Female"))) |>
    select(participant_id, visit, sex, {{ cols }}) |>
    tidyr::pivot_longer(cols = {{ cols }}, names_to = "what", values_to = "value") |>
    mutate(visit = paste0("yr-", visit) |> factor()) |>
    ggplot(aes(x = visit, y = value, fill = sex)) +
    geom_boxplot(col = "black") +
    facet_grid(~ what, scales = "free") +
    coord_flip() +
    scale_sex() +
    theme_workshop()
}

plot_boxplots(data, height:waist)
```

`stop()` halts the function with your message. The error names the function,
says what is wrong, and says what to do about it — the three things a good error
message needs.

## Going further

- [R for Data Science, chapter 25: Functions](https://r4ds.hadley.nz/functions.html)
  — vector, data frame and plot functions, at book length
- [Programming with dplyr](https://dplyr.tidyverse.org/articles/programming.html)
  — everything about `{{ }}`, `.data` and `...`

Back to [Module 2](module2.qmd).
