---
title: "Deep-dive · Data Types, Coercion and Factors"
subtitle: "Optional companion to Module 1"
---

Optional. Nothing in Module 2 assumes it. Read it when a column turns out to be
the wrong type and you want to know why.

```{r}
#| label: setup
library(readr)
library(dplyr)

abcd <- read_csv("data/abcd-synthetic.csv")
```

## The types, properly

```{r}
#| label: types
class(10.5)        # numeric  (specifically "double")
class(10L)         # integer  -- the L suffix forces a whole number
class("baseline")  # character
class(TRUE)        # logical
class(NA)          # logical, oddly -- see below
```

`class()` tells you what something is. `is.numeric()`, `is.character()` and
friends ask yes/no questions, which is more useful inside code:

```{r}
#| label: is-functions
is.numeric(abcd$age_months)
is.character(abcd$visit)
```

## Coercion, and why your numbers became text

R will not let a vector hold mixed types. When you try, it promotes everything to
the most accommodating type, following this one-way street:

```
logical  →  integer  →  double  →  character
```

Each step can represent everything to its left, so nothing is lost — but the
result may not be what you wanted:

```{r}
#| label: coercion-chain
c(TRUE, 1L)          # logical joins integer   -> integer
c(1L, 2.5)           # integer joins double    -> double
c(2.5, "hello")      # double joins character  -> character
```

That last one is the one that bites. A single stray text value anywhere in a
spreadsheet column turns the entire column into text:

```{r}
#| label: coercion-trap
ages <- c(118, 121, "not recorded", 130)
class(ages)
```

And once it is text, arithmetic fails:

```{r}
#| label: coercion-error
#| error: true
mean(ages)
```

In real data the culprit is usually a placeholder someone typed into a numeric
column — `"n/a"`, `"missing"`, `"999"`, `"<1"`, or a stray footnote marker.

### Fixing it

`as.numeric()` converts back. Anything that is not a number becomes `NA`, with a
warning telling you it happened:

```{r}
#| label: as-numeric
#| warning: true
ages_fixed <- as.numeric(ages)
ages_fixed
mean(ages_fixed, na.rm = TRUE)
```

That warning — `NAs introduced by coercion` — is useful, not noise. It is R saying
"I threw something away". If you see it unexpectedly, go and look at what got
discarded before moving on.

Better still, tell `read_csv()` about the placeholders up front so the column
arrives numeric in the first place:

```r
read_csv("data/my-file.csv", na = c("", "NA", "n/a", "missing", "999"))
```

## `NA` in more detail

`NA` is not zero, and not an empty string. It means *unknown*, and R propagates
that unknown-ness through everything it touches:

```{r}
#| label: na-propagation
NA + 1
NA > 5
NA == NA     # not TRUE -- two unknowns may or may not be equal
```

That last line is why you cannot test for missingness with `==`:

```{r}
#| label: na-test
x <- c(1, NA, 3)
x == NA        # useless
is.na(x)       # this is the way
```

There is one exception worth knowing, because it looks inconsistent until you see
the logic:

```{r}
#| label: na-logical
NA | TRUE      # TRUE  -- "unknown OR true" is true whatever the unknown is
NA & FALSE     # FALSE -- "unknown AND false" is false whatever the unknown is
```

## Factors

A **factor** is a categorical variable with a fixed, known set of possible values,
called its **levels**. It looks like text and behaves like a number underneath.

```{r}
#| label: factor-basic
visit_f <- factor(abcd$visit)
levels(visit_f)
```

Two reasons to care.

### Reason 1: controlling order

`factor()` sorts levels alphabetically by default. Sometimes that is wrong:

```{r}
#| label: factor-alpha
levels(factor(c("baseline", "year_1", "year_2")))
```

Here alphabetical happens to be correct — `b` before `y`. Now watch it fail with
more realistic visit labels:

```{r}
#| label: factor-alpha-fail
levels(factor(c("baseline", "2_year", "1_year")))
```

`"1_year"` sorts before `"2_year"` before `"baseline"`, so a plot would show the
timeline backwards. Set the order explicitly:

```{r}
#| label: factor-order
visits <- c("baseline", "2_year", "1_year")
factor(visits, levels = c("baseline", "1_year", "2_year")) |> levels()
```

This is *the* reason factors matter in practice: **ggplot2 and `group_by()` follow
factor level order**, so setting levels is how you control the order of bars,
boxes and legend entries.

```{r}
#| label: factor-in-pipeline
abcd |> 
  mutate(visit = factor(visit, levels = c("baseline", "year_1", "year_2"))) |> 
  count(visit)
```

### Reason 2: statistical models care

In Module 3, when you fit a model with a categorical predictor, R uses the **first
level as the reference group** — every coefficient is expressed relative to it.
Changing which level is first changes how the output reads. `relevel()` picks the
reference:

```{r}
#| label: relevel
sex_f <- factor(abcd$sex)
levels(sex_f)
levels(relevel(sex_f, ref = "M"))
```

### The classic factor trap

Converting a factor straight to numeric gives you the *level codes*, not the
values:

```{r}
#| label: factor-trap
f <- factor(c("10", "20", "30"))
as.numeric(f)              # 1 2 3 -- the internal codes. Wrong.
as.numeric(as.character(f))  # 10 20 30 -- correct
```

Go via `as.character()` every time. This one has silently corrupted a great many
analyses.

## Checking a fresh dataset

A habit worth building. Before analysing anything:

```{r}
#| label: sanity-check
glimpse(abcd)

# Any column that should be numeric but arrived as character?
abcd |> summarise(across(everything(), ~ class(.x)[1]))
```

If a column you expect to be numeric shows up as `character`, stop and find the
offending value before doing anything else.

Back to [Module 1](module1.qmd).
