---
title: "Module 1: Foundations of R Programming"
subtitle: "Biplabendu [ Billu ] Das"
pagetitle: "Module 1: Foundations of R Programming | ABCD R Workshop"
description-meta: "Learn R from scratch: read data with readr, wrangle it with dplyr and the |> pipe, handle missing values, and build your first ggplot2 histograms and boxplots."
open-graph:
  description: "Learn R from scratch: read data with readr, wrangle it with dplyr and the |> pipe, handle missing values, and build your first ggplot2 plots."
twitter-card:
  description: "Learn R from scratch: read data with readr, wrangle it with dplyr and the |> pipe, handle missing values, and build your first ggplot2 plots."
editor_options: 
  chunk_output_type: console
---

::: {.callout-warning}
# Page Under Development
This page is under active development. The contents might change.
:::

This module is for someone who has **never used R**. By the end of two hours you
will have read a real dataset into R, pulled it apart to answer questions about
it, dealt with missing values, and made three plots.

We teach R the way most modern R is written: **tidyverse** style, using the `|>`
pipe. Along the way we point out the bits of plain "base R" notation you need in
order to *read* code you find on the internet.

::: {.callout-note}
## Before you start
Work through [Module 0](module0.qmd) first — R and RStudio installed, an
RStudio Project open, the packages installed, and `abcd-synthetic.csv` sitting in
your `data/` folder.

Following along live? Open the skeleton script
[`module1-live.R`](../scripts/module1-live.R) and type into that.
:::

## 1. How code gets run

Open your project, then **File → New File → R Script**. You now have an empty
Editor pane at the top-left. Type this into it:

```{r}
#| label: first-line
2 + 2
```

With your cursor on that line, press **Ctrl+Enter** (**Cmd+Enter** on macOS). The
line jumps to the Console and the answer appears. That is the whole loop: write in
the Editor, run with Ctrl+Enter, read the answer in the Console.

The `[1]` in front of the answer is R telling you "this is the first element of
the result". It is noise for now. Ignore it.

### Storing things: `<-`

An answer that only prints is not much use. To keep it, give it a name:

```{r}
#| label: assignment
age_years <- 10
age_years
```

`<-` is the **assignment arrow**, and it means "put the thing on the right into
the name on the left". RStudio has a shortcut for it: **Alt+-** (**Option+-** on
macOS).

You will also see `=` used for assignment in other people's code. It mostly works.
Use `<-` anyway — it is the convention in R, and `=` has a different job inside
function calls, which we are about to meet.

::: {.callout-tip}
## Comments
Anything after a `#` is ignored by R. Comments are for the human reading the code
later, which is usually you, three weeks from now, with no memory of what you were
doing.

```r
# Convert age to months for the analysis
age_months <- age_years * 12
```
:::

## 2. The base R notation you need to read code

Four pieces of notation. You will see all four in every R script you ever open,
including the tidyverse code in this workshop.

### Functions and their arguments

A **function** does a job. You call it by name, with round brackets, and you put
what it should work on inside:

```{r}
#| label: functions
sqrt(16)
round(3.14159, digits = 2)
```

`sqrt` and `round` are the function names. `16`, `3.14159` and `digits = 2` are
**arguments** — the inputs. Named arguments use `=` (this is why `<-` is used for
assignment: it keeps the two ideas visually separate).

To find out what a function does and what arguments it takes, put a `?` in front:

```r
?round
```

The documentation appears in the Help pane, bottom-right. It is dense and takes
practice to read. Scroll straight to **Examples** at the bottom.

### `c()` — making a vector

`c()` is short for *combine*. It glues several values into one thing:

```{r}
#| label: vectors
scores <- c(104, 98, 116, 91)
scores
```

That is a **vector** — an ordered set of values of the same type. Almost every
column of every dataset you touch is one.

Operations apply to the whole vector at once, which is one of the genuinely nice
things about R:

```{r}
#| label: vectorised
scores + 10
mean(scores)
```

### `$` — pulling one column out of a table

Once we have a table of data, `$` reaches in and grabs a single column by name:

```r
abcd$age_months
```

You will see `$` constantly in older code. In tidyverse style we mostly avoid it,
because the functions we use let us refer to columns by bare name. But you need to
recognise it.

### `library()` — switching a package on

In Module 0 you *installed* packages. Installing downloads them once, permanently.
**Loading** switches one on for the current session, and you do it every time you
start R:

```{r}
#| label: libraries
library(readr)    # reading data files
library(dplyr)    # transforming data
library(tidyr)    # tidying and reshaping
library(ggplot2)  # plots
```

Put these lines at the top of every script. If you ever get
`Error: could not find function "filter"`, the cause is almost always a missing
`library()` line.

::: {.callout-note}
## Exercise 1.1
Make a vector called `sleep` holding the values 9.5, 8, 10, 7.5. Then find its
mean, and its highest value.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-1-1
sleep <- c(9.5, 8, 10, 7.5)
mean(sleep)
max(sleep)
```
:::

## 3. Data types

Every value in R has a **type**, and R is fussy about it. The four you will meet:

```{r}
#| label: types
class(10.5)          # numeric  -- numbers
class("baseline")    # character -- text, always in quotes
class(TRUE)          # logical  -- TRUE / FALSE
class(factor("F"))   # factor   -- categories with a fixed set of levels
```

Two consequences worth knowing now.

**Text is not a number.** `"10"` and `10` are different things, and `"10" + 1` is
an error, not `11`.

**Mixing types in a vector forces everything to the safest common type.** Put a
number and some text together and the number quietly becomes text:

```{r}
#| label: coercion
mixed <- c(1, 2, "three")
class(mixed)
mixed
```

Notice the quotes around `"1"` and `"2"`. This is called **coercion**, and it is a
common cause of "why is my column text?" — usually one stray non-numeric value in
a spreadsheet.

::: {.callout-tip}
## Going deeper
Types, coercion, and factors in more detail — including why factors bite people
when they reorder plots — are on the
[data types deep-dive](module1-data-types.qmd). Not needed today.
:::

## 4. Reading in the data

Now the real thing. `read_csv()` from **readr** reads a comma-separated file:

```{r}
#| label: read-data
abcd <- read_csv("data/abcd-synthetic.csv")
```

That path is **relative** to your project folder: "look in `data/`, find
`abcd-synthetic.csv`". It works because you opened the RStudio Project in
[Module 0](module0.qmd#set-up-a-project-folder) — that is what makes the short path
possible.

::: {.callout-important}
## If you get `Error: 'data/abcd-synthetic.csv' does not exist`
Two things to check, in this order:

1. Is the file actually in the `data/` folder? Look in the Files pane, bottom-right.
2. Is your project open? The title bar at the top-right of RStudio should say
   **r-workshop**. If it says *Project: (None)*, open the `.Rproj` file and try
   again.
:::

The dataset is **synthetic**: generated by a script, no real participants. But it
is deliberately shaped like an ABCD release — one row per participant per visit,
with realistic gaps.

## 5. Looking at what you loaded

Never analyse a dataset you have not looked at. Four functions, in the order you
should use them:

```{r}
#| label: glimpse
glimpse(abcd)
```

`glimpse()` is the one worth remembering: every column, its type, and the first
few values, all on one screen.

```{r}
#| label: inspect
dim(abcd)     # rows, then columns
names(abcd)   # column names
```

```{r}
#| label: head
head(abcd)    # the first 6 rows
```

```{r}
#| label: summary
summary(abcd)
```

`summary()` earns its keep for one reason: the `NA's` count at the bottom of a
column. That is your first look at missing data, and we come back to it in
[section 8](#missing-data).

Reading the columns:

| Column | What it is |
|:--|:--|
| `subject_id` | Participant identifier. **Repeats** — one row per visit. |
| `visit` | `baseline`, `year_1`, or `year_2` |
| `site` | Study site, `S01`–`S05` |
| `age_months` | Age at that visit, in months |
| `sex` | `F` or `M` |
| `total_problems` | A behaviour questionnaire total. Higher = more reported problems. |
| `cognition_score` | A cognition composite, centred near 100 |
| `screen_time_hrs` | Self-reported screen time, hours/day |
| `sleep_hrs` | Self-reported sleep, hours/night |

That `subject_id` repeats is the defining feature of a **longitudinal** dataset,
and it is the thing that makes ABCD data both powerful and easy to get wrong.
Module 3 deals with it properly.

::: {.callout-note}
## Exercise 1.2
How many rows and columns does `abcd` have? How many rows would you expect if
every one of the 200 participants attended all three visits — and does the number
match?
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-1-2
dim(abcd)
200 * 3
```

Fewer rows than 600. Some participants did not come back for a later visit —
**attrition**, and it is entirely normal in a real longitudinal study. Those
visits are missing as *absent rows*, not as `NA` values, which is a distinction
that matters later.
:::

## 6. The pipe: `|>`

Here is the idea that makes the rest of the workshop readable.

Say you want the mean of the cognition scores. Written as nested function calls,
you read it inside-out, right to left:

```{r}
#| label: nested
mean(abcd$cognition_score, na.rm = TRUE)
```

Fine for one step. But real analysis is five or six steps, and nesting becomes
unreadable. The **pipe** `|>` takes the thing on its left and passes it as the
first argument to the function on its right:

```{r}
#| label: pipe-simple
abcd$cognition_score |> mean(na.rm = TRUE)
```

Read `|>` out loud as **"and then"**. It lets you write a sequence in the order
you actually think about it:

> take the data, **and then** keep the baseline visits, **and then** average the
> cognition scores

The shortcut in RStudio is **Ctrl+Shift+M** (**Cmd+Shift+M**).

::: {.callout-tip}
## You will also see `%>%`
Older tidyverse code — which is to say most of the tutorials, Stack Overflow
answers and lab scripts you will ever meet — uses `%>%` instead. It comes from a
package called **magrittr** and does essentially the same job.

`|>` is newer, built into R itself, and needs no package. We use `|>` throughout.
When you see `%>%`, read it as `|>` and carry on.
:::

## 7. The five verbs that do the work

Almost all data wrangling is these five **dplyr** functions. Every one takes the
data as its first argument and returns a new table — which is exactly what makes
them pipeable.

### `filter()` — keep some rows

```{r}
#| label: filter
abcd |> filter(visit == "baseline")
```

Note `==`, two equals signs, for "is equal to". A single `=` is assignment, and
using it here is a common early error.

Several conditions, comma-separated, means *all* must hold:

```{r}
#| label: filter-two
abcd |> filter(visit == "baseline", sex == "F")
```

The comparisons available: `==` equal, `!=` not equal, `>` `<` `>=` `<=`, and
`%in%` for "is one of":

```{r}
#| label: filter-in
abcd |> filter(site %in% c("S01", "S02"))
```

### `select()` — keep some columns

```{r}
#| label: select
abcd |> select(subject_id, visit, cognition_score)
```

### `arrange()` — sort the rows

```{r}
#| label: arrange
abcd |> arrange(desc(cognition_score))
```

`desc()` sorts high-to-low; leave it off for low-to-high.

### `mutate()` — add or change a column

```{r}
#| label: mutate
abcd |> 
  mutate(age_years = age_months / 12) |> 
  select(subject_id, visit, age_months, age_years)
```

Notice the pipeline: add a column, **and then** pick the columns to show. This is
where the pipe starts paying off.

### `summarise()` — collapse to a single row

```{r}
#| label: summarise
abcd |> 
  summarise(
    n            = n(),
    mean_cog     = mean(cognition_score, na.rm = TRUE),
    mean_sleep   = mean(sleep_hrs, na.rm = TRUE)
  )
```

`n()` counts rows. On its own `summarise()` is mildly interesting; paired with
`group_by()` it becomes the workhorse of descriptive statistics:

```{r}
#| label: group-by
abcd |> 
  group_by(visit) |> 
  summarise(
    n        = n(),
    mean_cog = mean(cognition_score, na.rm = TRUE),
    mean_age = mean(age_months)
  )
```

One row per group. Read it as: split by `visit`, **and then** compute these
numbers within each split.

For the special case of just counting, `count()` is shorter:

```{r}
#| label: count
abcd |> count(visit)
abcd |> count(site, sex)
```

::: {.callout-note}
## Exercise 1.3
Using one pipeline: keep only the `year_2` visits, then report the mean
`screen_time_hrs` separately for each `sex`.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-1-3
abcd |> 
  filter(visit == "year_2") |> 
  group_by(sex) |> 
  summarise(
    n          = n(),
    mean_screen = mean(screen_time_hrs, na.rm = TRUE)
  )
```
:::

::: {.callout-note}
## Exercise 1.4
Add a column `sleep_deficit` that is `9 - sleep_hrs` (roughly, hours short of the
recommended nine). Then find the 5 participant-visits with the largest deficit,
showing only `subject_id`, `visit` and `sleep_deficit`.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-1-4
abcd |> 
  mutate(sleep_deficit = 9 - sleep_hrs) |> 
  arrange(desc(sleep_deficit)) |> 
  select(subject_id, visit, sleep_deficit) |> 
  head(5)
```
:::

## 8. Missing data {#missing-data}

`NA` means "not available" — the value is absent. Real datasets are full of them,
ABCD very much included, and R deliberately refuses to guess.

Watch what happens if you ignore them:

```{r}
#| label: na-trap
mean(abcd$cognition_score)
```

`NA`. Not a bug: R is telling you that the average of a set of numbers, one of
which is unknown, is itself unknown. To say "skip the missing ones and average the
rest", use `na.rm = TRUE`:

```{r}
#| label: na-rm
mean(abcd$cognition_score, na.rm = TRUE)
```

This is why `na.rm = TRUE` has been scattered through the code above.

### How much is missing?

`is.na()` gives `TRUE`/`FALSE` per value. Since R counts `TRUE` as 1, summing it
counts the missing values:

```{r}
#| label: count-na
abcd |> 
  summarise(
    missing_cog    = sum(is.na(cognition_score)),
    missing_probs  = sum(is.na(total_problems)),
    missing_screen = sum(is.na(screen_time_hrs))
  )
```

Better still, look at whether missingness is spread evenly — because if it is not,
dropping those rows will bias your results:

```{r}
#| label: na-by-visit
abcd |> 
  group_by(visit) |> 
  summarise(
    n           = n(),
    missing_cog = sum(is.na(cognition_score)),
    pct_missing = round(100 * mean(is.na(cognition_score)), 1)
  )
```

### Dropping rows

`drop_na()` from **tidyr** removes rows with missing values in the columns you
name:

```{r}
#| label: drop-na
abcd_cog <- abcd |> drop_na(cognition_score)
nrow(abcd)
nrow(abcd_cog)
```

::: {.callout-warning}
## Dropping is a decision, not a cleanup step
`drop_na()` is easy to type and easy to over-use. Every row you drop is a real
participant whose data you have chosen to exclude, and if the people with missing
scores differ systematically from those without, your results shift.

For today, dropping is fine — we are learning the mechanics. **Module 3** covers
what to actually do about missing data in a longitudinal study, which is a
genuinely hard problem and not one `drop_na()` solves.
:::

## 9. Your first plots

**ggplot2** builds a plot in layers, joined by `+`. Two pieces are always present:

- `aes()` — the **aes**thetic mapping: which column goes on which axis
- a `geom_*()` — the kind of mark to draw

### A histogram: the shape of one variable

```{r}
#| label: fig-histogram
#| fig-cap: "Distribution of cognition scores across all visits"
#| fig-alt: "Histogram of cognition scores, roughly bell-shaped and centred near 100."
#| warning: false
abcd |> 
  ggplot(aes(x = cognition_score)) +
  geom_histogram(binwidth = 5, fill = "#62272D", colour = "white") +
  labs(
    title = "Cognition scores are roughly bell-shaped",
    x     = "Cognition score",
    y     = "Number of participant-visits"
  )
```

Note the pipe carries the data *in*, and then layers are added with `+`. Mixing
the two up — writing `|>` between layers — is the most common ggplot2 error there
is.

### A boxplot: comparing groups

```{r}
#| label: fig-boxplot
#| fig-cap: "Total problems by visit"
#| fig-alt: "Boxplots of total problem scores at baseline, year 1 and year 2, with slightly lower medians at later visits."
#| warning: false
abcd |> 
  ggplot(aes(x = visit, y = total_problems)) +
  geom_boxplot(fill = "#FDBF6F") +
  labs(
    title = "Reported problems drift down slightly across visits",
    x     = "Visit",
    y     = "Total problems"
  )
```

### A line chart: change over time

For a trend you first have to *compute* the trend — one number per visit — and
then plot it. Which is the `group_by()` + `summarise()` pattern from
[section 7](#the-five-verbs-that-do-the-work), piped straight into a plot:

```{r}
#| label: fig-line
#| fig-cap: "Mean cognition score at each visit"
#| fig-alt: "Line chart with points showing mean cognition score rising modestly from baseline through year 2."
abcd |> 
  group_by(visit) |> 
  summarise(mean_cog = mean(cognition_score, na.rm = TRUE)) |> 
  ggplot(aes(x = visit, y = mean_cog, group = 1)) +
  geom_line(colour = "#62272D", linewidth = 1) +
  geom_point(size = 3, colour = "#62272D") +
  labs(
    title = "Mean cognition score rises across visits",
    x     = "Visit",
    y     = "Mean cognition score"
  )
```

`group = 1` looks odd. `visit` is text, not a number, so ggplot2 does not know
which points to join up; `group = 1` says "these all belong to one line". You will
need it every time you draw a line over a categorical x-axis.

::: {.callout-note}
## Exercise 1.5
Draw a boxplot of `sleep_hrs` by `visit`. Then add `colour = sex` inside `aes()`
and see what changes.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-1-5
#| fig-alt: "Boxplots of sleep hours by visit, split by sex."
#| warning: false
abcd |> 
  ggplot(aes(x = visit, y = sleep_hrs, colour = sex)) +
  geom_boxplot() +
  labs(x = "Visit", y = "Sleep (hours/night)")
```

Mapping a column to `colour` splits each box in two and adds a legend
automatically. That is the pattern for "show me this broken down by group".
:::

::: {.callout-tip}
## Going deeper
More geoms, faceting into small multiples, and how to make a figure that would
pass in a paper are in the [plot gallery](module1-plot-gallery.qmd). Module 2 is
devoted to this.
:::

## 10. Saving your work

A result that only exists in your Console is not a result. `write_csv()` puts a
table on disk:

```{r}
#| label: write-out
visit_summary <- abcd |> 
  group_by(visit) |> 
  summarise(
    n        = n(),
    mean_cog = mean(cognition_score, na.rm = TRUE),
    mean_probs = mean(total_problems, na.rm = TRUE)
  )

write_csv(visit_summary, "output/visit-summary.csv")
```

And `ggsave()` saves the most recent plot:

```r
ggsave("output/cognition-by-visit.png", width = 6, height = 4, dpi = 300)
```

This is why Module 0 had you make an `output/` folder. Keeping saved results
separate from raw data means you can always delete everything in `output/` and
rebuild it by re-running the script — which is a decent working definition of a
reproducible analysis.

## Putting it together

Everything from today, as one pipeline — read, clean, summarise, save:

```{r}
#| label: full-pipeline
abcd |> 
  filter(visit == "baseline") |> 
  drop_na(cognition_score) |> 
  mutate(age_years = age_months / 12) |> 
  group_by(sex) |> 
  summarise(
    n            = n(),
    mean_age     = round(mean(age_years), 1),
    mean_cog     = round(mean(cognition_score), 1),
    mean_screen  = round(mean(screen_time_hrs, na.rm = TRUE), 1)
  )
```

Read it as one sentence: take the data, keep baseline visits, drop rows missing a
cognition score, add an age-in-years column, split by sex, and report these four
numbers.

Two hours ago that was an opaque wall of symbols.

## What you can now do

- Run code from a script, and store results with `<-`
- Read function calls, `c()`, `$` and `library()` when you meet them
- Load a CSV and inspect it before trusting it
- Chain steps with `|>`
- `filter()`, `select()`, `arrange()`, `mutate()`, `group_by()` + `summarise()`
- Spot missing data, count it, and decide about it deliberately
- Build histograms, boxplots and line charts, and save them

## Going further

Optional, and none of it is assumed in Module 2:

- [Data types deep-dive](module1-data-types.qmd) — coercion and factors in detail
- [Plot gallery](module1-plot-gallery.qmd) — more geoms and faceting
- [R for Data Science](https://r4ds.hadley.nz), chapters 1–4 — the same ground, at
  book length
- The [dplyr cheatsheet](https://rstudio.github.io/cheatsheets/data-transformation.pdf) —
  one page, keep it next to you

Next: [Module 2 · Data Exploration & Visualization →](module2.qmd) on Sep 24, where
the pipelines get longer and the plots get good.
