---
title: "Module 2 · Data Exploration, Visualization & Descriptive Statistics"
subtitle: "Thu Sep 24, 2026 · noon–2pm PT"
pagetitle: "Module 2: Data Visualization with ggplot2 | ABCD R Workshop"
description-meta: "Build tidy data pipelines in R and create publication-ready figures with ggplot2. Module 2 of a four-part R and statistics workshop for ABCD researchers."
open-graph:
  description: "Tidy data pipelines and publication-ready ggplot2 figures. Module 2 of the ABCD R and statistics workshop series."
twitter-card:
  description: "Tidy data pipelines and publication-ready ggplot2 figures. Module 2 of the ABCD R and statistics workshop series."
editor_options:
  chunk_output_type: console
---

**Designed for** someone with a basic knowledge of R and RStudio — which
[Module 1](module1.qmd) provides.

Module 1 used a small, tidy teaching file with plain English column names. Today
we move to data shaped like a real ABCD release: cryptic variable names, values
stored as numeric codes, and a lot of missing cells. The work of the session is
getting from that raw file to two figures you could put on a poster:

- boxplots of height, weight and waist circumference at every visit, split by sex
- the mean weekday screen time at each age, with standard-error bars

Along the way we build longer **data pipelines** — select, rename, recode, join,
reshape — and learn how **ggplot2** figures are put together layer by layer.

::: {.callout-note}
## Before you start
Download the three files from the [dataset page](data.qmd#get-the-files) —
`data.RDS`, `data_dictionary.csv` and `data_dictionary_levels.csv` — into your
project's `data/` folder, next to `abcd-synthetic.csv` from Module 1.

Following along live? The code from the session is in
[`module2-live.R`](../scripts/module2-live.R). Every chunk on this page is taken
from it, in the same order.
:::

## Recap

Everything today is built from pieces you met in [Module 1](module1.qmd):

- `|>` — "and then": hands the result on the left to the function on the right
- `select()` keeps columns, `filter()` keeps rows, `mutate()` adds or changes columns
- `group_by()` + a summary function computes one row per group
- `count()` tabulates
- `ggplot(aes(...)) + geom_*()` — a plot is a mapping plus layers, joined by `+`

If any of those is unfamiliar, Module 1 sections 6, 7 and 9 are the place to
catch up.

## 0. Load libraries

```{r}
#| label: libraries
library(dplyr)
library(ggplot2)
```

Only two `library()` calls, yet the code below also uses functions from
**readr**, **stringr** and **tidyr**. It reaches them with `package::function()`
notation instead — `readr::read_csv()` means "the `read_csv()` function from the
readr package".

That notation does two things. It uses a function without switching the whole
package on, and it makes it obvious to the reader where a function comes from.
It is a good habit for packages you only need once or twice in a script. The
packages still have to be *installed* (Module 0 did that); `::` just skips the
*loading*.

## 1. Reading files into R

### CSV

```{r}
#| label: read-csv
# - CSV

readr::read_csv("data/abcd-synthetic.csv") |>
  glimpse()
```

This is the Module 1 file, read and glimpsed but not stored — there is no `<-`,
so the result is printed and then thrown away. It is here as a reminder of what
a CSV gives you: plain columns of numbers and text.

### RDS

```{r}
#| label: read-rds
# - RDS (R Data Set)

data <- readRDS("data/data.RDS") |>
  glimpse()
```

`.RDS` is R's own file format. It stores an R object exactly as it was, so
factors stay factors (`<fct>`) and the visit timestamp stays a date-time
(`<dttm>`). A CSV would flatten both into text or numbers, and you would rebuild
them by hand every time.

Why does `data` end up holding the data, when the last step of the pipe is
`glimpse()`? Because `glimpse()` prints a preview and then **passes its input
through unchanged**. That makes it safe to drop into the middle of any pipeline
to peek at what is flowing through — a trick used throughout this session.

2,295 rows and 53 columns: 500 participants, each with up to five annual visits.
What the column names mean, and how to read them, is on the
[dataset page](data.qmd#how-abcd-names-its-variables) — keep it open in another
tab.

## 2. Select columns of interest

Fifty-three columns is too many to think about. The first step is to cut down to
the ones the analysis needs, and give the important ones names a human can type.

```{r}
#| label: select
sub <- data |>
  select(
    # by position
    1:2,
    # by name, simultaneously rename
    age = ab_g_dyn__visit_age,
    site = ab_g_dyn__design_site,
    income = ab_g_dyn__cohort_income__hhold__3lvl,
    sex = ab_g_stc__cohort_sex,
    ethn = ab_g_stc__cohort_ethn,
    # by pattern
    matches("_anthr_"),
    matches("_screen_"),
    matches("_cbcl_")
  ) |>
  rename(
    height = ph_y_anthr__height_mean,
    weight = ph_y_anthr__weight_mean,
    waist = ph_y_anthr__waist_001
  ) |>
  as_tibble() |>
  glimpse()
```

`select()` can pick columns in three different ways, and this one call uses all
three:

- **By position.** `1:2` means "columns 1 through 2" — here `participant_id` and
  `session_id`. Quick, but fragile: if the column order ever changes, so does
  what you get.
- **By name, with a rename.** `age = ab_g_dyn__visit_age` keeps the column *and*
  renames it, `new_name = old_name`. The same pattern works in `rename()`.
- **By pattern.** `matches("_anthr_")` keeps every column whose name contains
  `_anthr_` — all three anthropometric measures at once. This is where ABCD's
  structured names pay off: one short pattern selects a whole instrument.

`rename()` then does the same `new = old` renaming without dropping anything.
The three anthropometric columns came in via `matches()`, so they are renamed
here rather than inside `select()`.

`as_tibble()` makes sure the result is a **tibble**, the tidyverse flavour of a
data frame, which prints more compactly and more safely than a plain
`data.frame`.

::: {.callout-tip}
## Going deeper
`starts_with()`, `ends_with()`, `contains()` and `where()` — the other ways to
pick columns — are on the [wrangling deep-dive](module2-wrangling.qmd#choosing-columns).
:::

## 3. Format data

A dataset fresh from a release is rarely ready to analyse. Four checks, every time:

1. **Check data types.** Is every column the type you expect — numbers as
   numbers, categories as factors?
2. **Rename factor levels.** Coded values (`1`, `2`) need their labels
   (`Male`, `Female`) before they are useful in a table or a plot legend.
3. **Look for implausible values.** Data-entry errors and impossible values
   hide in every real dataset.
4. **Handle missingness.** Know where the `NA`s are before you decide what to do
   about them.

### The levels dictionary

The labels for coded variables do not live in the data file. They live in
`data_dictionary_levels.csv` — one row per code, per variable:

```{r}
#| label: levels-dict
levels_dict <- readr::read_csv("data/data_dictionary_levels.csv") |>
  glimpse()
```

`site` is stored as a number from 1 to 22. To turn those into site names, pull
out just the site rows:

```{r}
#| label: site-levels
site_levels <- levels_dict |>
  filter(
    name == "ab_g_dyn__design_site"
  ) |>
  select(
    value,
    site_label = label
  )
```

`site_levels` is now a small two-column lookup table: `value` (the code) and
`site_label` (what it means). The `select()` renames `label` on the way out, so
it cannot clash with anything later.

### Recoding and joining

```{r}
#| label: format-data
sub2 <- sub |>
  glimpse() |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x)
    )
  ) |>
  mutate(
    sex = case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female",
      .default = NA
    ),
    visit = stringr::str_replace(session_id, "ses-", "") |>
      readr::parse_number() |>
      factor(),
    site = as.character(site)
  ) |>
  left_join(
    site_levels |>
      mutate(
        value = as.character(value)
      ),
    join_by(site == value)
  ) |>
  glimpse()
```

A lot happens in one pipeline, so take it a step at a time.

**`across(where(is.numeric), ~ round(.x))`** — "for every column that is
numeric, replace it with a rounded version of itself". `across()` applies one
function to many columns; `where(is.numeric)` picks which ones; `~ round(.x)` is
the function, with `.x` standing for "the current column". The point is to turn
age (`10.2`, `10.7`, …) into whole years, which the plot in Q2 needs.

Be aware that it rounds *every* numeric column — height, weight, screen time and
the CBCL scores too. That is fine for the plots today, but in a real analysis you
would usually round only the column you meant to, e.g. `mutate(age = round(age))`.

**`case_when()`** — recodes by condition. Each line is `condition ~ value`, tried
in order; `.default = NA` catches anything that matched no line. `sex` is a
factor whose levels are `"1"` and `"2"`, and `sex == 1` compares against those
levels. The result is plain text, `"Male"` or `"Female"`.

**Building `visit`** — `session_id` looks like `ses-02A`. `str_replace()`
removes the `ses-` prefix, leaving `02A`; `parse_number()` pulls out the number,
`2`; `factor()` makes it a category. So `visit` runs `0` to `4` — baseline to
year-4.

**`site = as.character(site)`** — converting to text ahead of the join. See the
next paragraph.

**`left_join()`** — this is new, and it is the most important step here. A
**join** matches rows in two tables by a shared key and glues on the columns from
the second table. `join_by(site == value)` says "match `site` in our data against
`value` in the lookup". **Left** join means every row of our data is kept,
whether or not it finds a match.

Joins only match when the keys are the *same type*. `site` in the data is a
factor; `value` in the lookup is a number. Converting both to character first —
`as.character(site)` above, `as.character(value)` inside the join — puts them on
common ground.

::: {.callout-tip}
## Going deeper
Other join types, what happens when the key types do not match, and how to label
*all thirteen* coded variables in one go, are on the
[wrangling deep-dive](module2-wrangling.qmd#joins).
:::

After any join, check it did what you expected:

```{r}
#| label: check-join
sub2 |>
  count(site, site_label)
```

Every code has a name next to it, and no `site_label` is `NA`. If a join
silently fails to match, this is where you see it.

### Implausible values

The third check on the list — looking for values that cannot be right.
`summary()` on the columns you plan to use is the fastest way:

```{r}
#| label: implausible
sub2 |>
  select(height, weight, waist, matches("_screen_")) |>
  summary()
```

Read the `Min.` and `Max.` rows. A weekend screen-time total of 150 hours is
impossible, however much a 13-year-old might want it to be, and a waist
circumference of 12 inches is not a plausible child. The medians look sensible.
It is a handful of extreme values that do not.

::: {.callout-warning}
## Finding is not fixing
Spotting implausible values is a descriptive job, and it belongs here. Deciding
what to *do* about them — drop them, cap them, or treat them as missing — is an
analysis decision with consequences, and it gets proper treatment in
[Module 3](module3.qmd). For now, keep an eye out for them in the plots below:
boxplots are very good at showing them.
:::

## Q1. Height, Weight, and Waist

The first question: how do body measurements change across visits, and do they
differ by sex? Start with one measure, waist circumference:

```{r}
#| label: fig-waist
#| fig-cap: "Waist circumference by visit and sex"
#| fig-alt: "Horizontal boxplots of waist circumference for visits 0 to 4, with orange boxes for males and grey boxes for females; waist increases slightly across visits."
#| warning: false
sub2 |>
  glimpse() |>
  # tidyr::drop_na(height) |>
  mutate(
    sex = factor(
      sex,
      levels = c("Male", "Female")
    ),
    visit = stringr::str_replace(session_id, "ses-", "") |>
      readr::parse_number() |>
      factor()
  ) |>
  glimpse() |>
  ggplot(
    aes(
      x = visit,
      # x = session_id,
      y = waist,
      fill = sex,
      col = sex
    )
  ) +
  theme_bw(20) +
  # geom_hline(yintercept = 60, colour = "grey60") +
  # geom_vline(xintercept = 12.5, colour = "red") +
  # geom_point()
  # geom_violin(col = "black") +
  geom_boxplot(col = "black") +
  # geom_jitter(alpha = 0.5, size = 0.2) +
  # scale_fill_viridis_d() +
  scale_fill_manual(values = c("orange", "grey60")) +
  scale_color_viridis_d() +
  coord_flip() +
  theme(
    legend.position = "top"
  )
```

Before the plot, the `mutate()` does two things:

- **`sex = factor(sex, levels = c("Male", "Female"))`** fixes the order of the
  categories. Left alone, R sorts text alphabetically, so `Female` would come
  first. The order of the levels is the order of the legend and of the colours.
- **`visit = ...`** rebuilds `visit` from `session_id`. `sub2` already has an
  identical `visit` column from section 3, so this line is redundant — harmless,
  but you could delete it and get the same plot.

Then the plot, one piece at a time:

- **`aes(x = visit, y = waist, fill = sex, col = sex)`** — visit on one axis,
  waist on the other, and sex mapped to two things: **`fill`**, the inside colour
  of a shape, and **`col`** (colour), its outline.
- **`theme_bw(20)`** — a clean black-and-white theme, with a base font size of
  20 so the text is readable when shared on screen or put on a poster.
- **`geom_boxplot(col = "black")`** — draw the boxes. `col = "black"` *outside*
  `aes()` sets the outline to one fixed colour, overriding the mapping. Inside
  `aes()` maps a column; outside sets a constant.
- **`scale_fill_manual(values = c("orange", "grey60"))`** — choose the fill colours
  yourself, in the order of the factor levels: Male orange, Female grey.
- **`scale_color_viridis_d()`** — a colour-blind-friendly palette for the outline
  mapping. It has no visible effect here, because `col = "black"` in the geom
  wins; it shows up when you switch geoms (try `geom_point()`).
- **`coord_flip()`** — swap the axes so the boxes run horizontally. Long labels
  and many categories read more easily this way.
- **`theme(legend.position = "top")`** — move the legend from the right to the
  top, giving the plot more width.

::: {.callout-note}
## Try it: uncomment a line
The commented-out lines are alternatives from the live session. Remove the `#`
from one at a time and re-run:

- `geom_violin(col = "black")` — the full shape of the distribution, not just
  its quartiles
- `geom_jitter(alpha = 0.5, size = 0.2)` — every individual measurement as a
  faint point, on top of the boxes
- `scale_fill_viridis_d()` — replace the manual colours with the viridis palette
  (comment out `scale_fill_manual()` first, or R warns that it is replacing it)
- `geom_hline(...)`, `geom_vline(...)` — reference lines. Note what `coord_flip()`
  does to them: a "horizontal" line ends up vertical.
- `x = session_id` in place of `x = visit` — the raw visit codes instead of 0–4
- `tidyr::drop_na(height)` — drop the rows missing height before plotting
:::

### All three at once: reshape, then facet

Doing that three times, once per measure, is repetitive. Better to put all three
side by side in one figure. That takes one new idea: **reshaping**.

```{r}
#| label: fig-anthro-facet
#| fig-cap: "Height, waist and weight by visit and sex"
#| fig-alt: "Three panels of horizontal boxplots — height, waist and weight — for visits yr-0 to yr-4, orange for males and grey for females; each measure increases across visits, with males taller and heavier by year 4."
#| fig-width: 12
#| fig-height: 6
#| warning: false
sub2 |>
  mutate(
    sex = factor(
      sex,
      levels = c("Male", "Female")
    )
  ) |>
  select(
    participant_id,
    visit,
    sex,
    height,
    weight,
    waist
  ) |>
  tidyr::pivot_longer(
    cols = height:waist,
    names_to = "what",
    values_to = "value"
  ) |>
  mutate(
    visit = paste0("yr-", visit) |> factor()
  ) |>
  ggplot(
    aes(
      x = visit,
      y = value,
      fill = sex,
      col = sex
    )
  ) +
  geom_boxplot(col = "black") +
  facet_grid(~ what, scales = "free") +
  coord_flip() +
  theme_bw(20) +
  scale_fill_manual(values = c("orange", "grey60")) +
  scale_color_viridis_d() +
  # labs(
  #   x = "",
  #   y = "",
  #   title = "Title here",
  #   subtitle = "sub-title goes here",
  #   caption = "some additional info here"
  # ) +
  theme(
    legend.position = "top"
  )
```

The data so far is **wide**: one row per participant-visit, with height, weight
and waist in three separate columns. ggplot2 can only facet by a *column*, so it
needs the data **long**: one row per participant-visit-*measure*, with a column
saying which measure it is.

`tidyr::pivot_longer()` does that conversion:

- `cols = height:waist` — the columns to stack (`:` means "height through waist",
  in column order)
- `names_to = "what"` — a new column holding the old column *names*
- `values_to = "value"` — a new column holding the *numbers*

So each participant-visit row becomes three rows, `what` says `height`, `weight`
or `waist`, and `value` holds the measurement. To see it for yourself, run the
pipeline up to and including `pivot_longer()` and then `glimpse()`.

`paste0("yr-", visit)` glues a prefix onto the visit number (`0` → `yr-0`) so the
axis labels mean something without a caption.

`facet_grid(~ what, scales = "free")` draws one panel per value of `what`. The
`~ what` puts them in columns, side by side. `scales = "free"` lets each panel
use its own value range — essential here, because height is in inches, weight
in pounds, and on a shared axis the height boxes would be squashed into a sliver.
The panels come out in alphabetical order, height–waist–weight.

The commented-out `labs()` block is how you would add axis titles, a title, a
subtitle and a caption for a figure headed for a paper or poster. Fill it in and
uncomment it.

::: {.callout-tip}
## Going deeper
Fill vs colour for every geom, choosing palettes, `facet_grid()` vs
`facet_wrap()`, and saving a figure at print resolution with `ggsave()` are on
the [ggplot2 anatomy deep-dive](module2-ggplot-anatomy.qmd).
:::

## Exercise 1: boxplots of the `_screen_` variables

::: {.callout-note}
## Exercise 1 · ~15 min
`sub2` has four screen-time columns — the ones whose names contain `_screen_`:
youth- and parent-reported, weekday and weekend. Make the same faceted boxplot
for those four, one panel per screen-time measure.

Hint: copy the height/weight/waist pipeline and change the columns you select
and pivot. `matches("_screen_")` works inside `pivot_longer()` too.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-1
#| fig-alt: "Four panels of horizontal boxplots of screen time by visit and sex — parent-reported and youth-reported, weekday and weekend — with a few extreme outliers stretching the axes."
#| fig-width: 12
#| fig-height: 6
#| warning: false
sub2 |>
  mutate(
    sex = factor(
      sex,
      levels = c("Male", "Female")
    )
  ) |>
  select(
    participant_id,
    visit,
    sex,
    matches("_screen_")
  ) |>
  tidyr::pivot_longer(
    cols = matches("_screen_"),
    names_to = "what",
    values_to = "value"
  ) |>
  mutate(
    visit = paste0("yr-", visit) |> factor(),
    what = stringr::str_replace(what, "__screen__", "\n")
  ) |>
  ggplot(
    aes(
      x = visit,
      y = value,
      fill = sex
    )
  ) +
  geom_boxplot(col = "black") +
  facet_grid(~ what, scales = "free") +
  coord_flip() +
  theme_bw(20) +
  scale_fill_manual(values = c("orange", "grey60")) +
  theme(
    legend.position = "top"
  )
```

The one addition is `str_replace(what, "__screen__", "\n")`, which swaps the
middle of each long variable name for a line break (`\n`) so the panel labels
fit. Read `nt_p_yst` as the parent report and `nt_y_stq` as the youth's own.

Look at how far the axes stretch. Those long tails are the implausible values
from [section 3](#implausible-values) — 80 hours on a weekday — and they squash
the boxes towards zero. A boxplot is often the first place a data-quality
problem shows itself.

Notice too that youth report more screen time than their parents do, and
weekends more than weekdays.
:::

## Q2. Mean + Error bars

The second question: how does weekday screen time change with age, and does it
differ by sex? This time we plot a *summary* — the mean at each age — rather than
the whole distribution.

```{r}
#| label: fig-screen-mean-se
#| fig-cap: "Mean weekday screen time (youth report) by age and sex, ± 1 standard error"
#| fig-alt: "Points joined by lines showing mean weekday screen time rising from about 4 hours a day at age 10 to about 8 to 10 hours at age 15, for females in orange and males in grey, with small error bars."
#| fig-width: 10
#| fig-height: 6
sub2 |>
  # summary()
  select(
    participant_id,
    visit,
    age,
    sex,
    screen_wkday = nt_y_stq__screen__wkdy_sum,
    screen_wknd = nt_y_stq__screen__wknd_sum,
    cbcl_int = mh_p_cbcl__synd__int_sum,
    cbcl_ext = mh_p_cbcl__synd__ext_sum
  ) |>
  # count(age)
  filter(
    age >= 10 & age <= 15,
    !is.na(screen_wkday),
    !is.na(screen_wknd),
    !is.na(cbcl_int),
    !is.na(cbcl_ext)
  ) |>
  # summary()
  mutate(
    age = factor(age)
  ) |>
  group_by(age, sex) |>
  reframe(
    n = n(),
    mean_screen_wkday = mean(screen_wkday),
    mean_screen_wknd = mean(screen_wknd),
    sd_wkday = sd(screen_wkday),
    sd_wknd = sd(screen_wknd)
  ) |>
  mutate(
    mean = mean_screen_wkday,
    sd = sd_wkday,
    se = sd / sqrt(n)
  ) |>
  ggplot(
    aes(x = age, y = mean, fill = sex, col = sex, group = sex)
  ) +
  # geom_ribbon(
  #   aes(
  #     ymin = mean - se,
  #     ymax =  mean + se
  #   ),
  #   alpha = 0.5
  # ) +
  # geom_line(
  #   size = 2,
  #   aes(col = sex)
  # ) +
  geom_errorbar(
     aes(
       ymin = mean - se,
       ymax =  mean + se
     ),
     position = position_dodge(width = 0.7),
     width = .3
  ) +
  geom_line(
    position = position_dodge(width = 0.7)
  ) +
  geom_point(
    position = position_dodge(width = 0.7),
    size = 7
  ) +
  theme_bw(20) +
  scale_color_manual(values = c("orange", "grey60")) +
  scale_fill_manual(values = c("orange", "grey60")) +
  theme(
    legend.position = "top"
  )
```

The longest pipeline so far. Taken in order:

**Select and rename.** Keep eight columns, and give the four measures short names.

**Filter.** `age >= 10 & age <= 15` keeps ages 10 to 15 — there are very few
participants at 8, 9, 16 or 17, and means from a handful of people are noisy.
Run the commented-out `count(age)` to see why. The four `!is.na(...)` conditions
drop rows missing any of the measures, so every mean below is computed on the
same people. (`!` means "not": `!is.na(x)` is "x is not missing".)

**`age = factor(age)`.** Age is a whole number after the rounding in section 3.
Making it a factor tells ggplot2 to treat each age as its own category on the
x-axis, rather than as a continuous number line.

**`group_by(age, sex)` + `reframe()`.** One row per age-and-sex combination,
holding the count `n()`, the mean and the **standard deviation** (`sd()`) of
weekday and weekend screen time. `reframe()` is a close relative of the
`summarise()` from Module 1. Here the two give exactly the same result — use
whichever you like. (`summarise()` would also print a message about the
grouping; `reframe()` always returns an ungrouped table.)

**`se = sd / sqrt(n)`.** The **standard error** of the mean. The `mutate()` first
copies the weekday columns into generic `mean` and `sd` columns — so to plot
weekend screen time instead, you change those two lines and nothing else.

**The plot.**

- `group = sex` tells `geom_line()` which points to join — one line per sex.
  Without it, ggplot2 cannot tell which points belong together on a categorical
  x-axis. It is the same job `group = 1` did in Module 1's line chart.
- `geom_errorbar(aes(ymin = mean - se, ymax = mean + se))` draws a bar from one
  standard error below the mean to one above. `width = .3` is the width of the
  caps.
- `position = position_dodge(width = 0.7)` nudges the two sexes apart
  horizontally so their points and bars do not sit on top of each other. It must
  be the **same** on every layer, or the bars, lines and points drift out of
  alignment.
- `geom_point(size = 7)` — the means themselves, drawn large.

::: {.callout-important}
## Why are the colours swapped?
In Q1, males were orange. Here, females are. This pipeline never sets the factor
order of `sex`, so R sorts the text alphabetically — `Female` before `Male` —
and `scale_*_manual()` hands out colours in that order.

To make the colours consistent across figures, add the same
`mutate(sex = factor(sex, levels = c("Male", "Female")))` step from Q1. Setting
the level order once, early in the pipeline, is the reliable fix.
:::

The commented-out `geom_ribbon()` + `geom_line()` pair is an alternative way to
show the same numbers: a shaded band for ± 1 SE instead of bars, the usual style
when x is a continuous variable. (If you uncomment it, change `size = 2` to
`linewidth = 2` — `size` for lines is deprecated in current ggplot2.)

### The numbers behind the plot

A plot of means is a picture of a table. Stop the same pipeline before
`ggplot()` to see that table:

```{r}
#| label: descriptive-table
sub2 |>
  select(
    age,
    sex,
    screen_wkday = nt_y_stq__screen__wkdy_sum
  ) |>
  filter(
    age >= 10 & age <= 15,
    !is.na(screen_wkday)
  ) |>
  group_by(age, sex) |>
  summarise(
    n    = n(),
    mean = mean(screen_wkday),
    sd   = sd(screen_wkday),
    se   = sd / sqrt(n),
    .groups = "drop"
  ) |>
  mutate(across(mean:se, ~ round(.x, 2)))
```

(The counts differ slightly from the plot's, because this version only requires
weekday screen time to be present, not all four measures.)

This is a **descriptive statistics** table, and it is often what a paper's
Table 1 is built from. Two numbers there describe spread, and they are easy to
confuse:

- **SD, standard deviation** — how much *individuals* vary around the mean. It
  describes the people, and it does not shrink as you collect more of them.
- **SE, standard error** — how precisely the *mean* is estimated. It is SD
  divided by √n, so it shrinks as n grows — and grows with SD. The widest error
  bar is for girls at 15: one of the smallest groups, and one of the most spread
  out.

::: {.callout-warning}
## Error bars describe, they do not test
Non-overlapping error bars are *not* a test of whether two groups differ, and
overlapping ones do not show that they are the same. Whether the gap between
girls and boys at 15 is more than noise is a question for a hypothesis test —
which is exactly what [Module 3](module3.qmd) is about.
:::

## Exercise 2: a different measure

::: {.callout-note}
## Exercise 2 · ~10 min
The Q2 pipeline already carries `cbcl_int`, the parent-reported CBCL
internalizing score. Make the mean ± SE plot for `cbcl_int` by age and sex
instead of screen time. Set the `sex` level order so that males are orange, as
in Q1.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: ex-2
#| fig-alt: "Mean CBCL internalizing score by age for males in orange and females in grey, with error bars; both rise from about 4 at age 10, and females are higher from age 11 onward, reaching about 7 at 15."
#| fig-width: 10
#| fig-height: 6
sub2 |>
  select(
    age,
    sex,
    cbcl_int = mh_p_cbcl__synd__int_sum
  ) |>
  filter(
    age >= 10 & age <= 15,
    !is.na(cbcl_int)
  ) |>
  mutate(
    age = factor(age),
    sex = factor(sex, levels = c("Male", "Female"))
  ) |>
  group_by(age, sex) |>
  reframe(
    n = n(),
    mean = mean(cbcl_int),
    sd = sd(cbcl_int),
    se = sd / sqrt(n)
  ) |>
  ggplot(
    aes(x = age, y = mean, fill = sex, col = sex, group = sex)
  ) +
  geom_errorbar(
    aes(
      ymin = mean - se,
      ymax = mean + se
    ),
    position = position_dodge(width = 0.7),
    width = .3
  ) +
  geom_line(
    position = position_dodge(width = 0.7)
  ) +
  geom_point(
    position = position_dodge(width = 0.7),
    size = 7
  ) +
  theme_bw(20) +
  scale_color_manual(values = c("orange", "grey60")) +
  scale_fill_manual(values = c("orange", "grey60")) +
  labs(
    x = "Age (years)",
    y = "CBCL internalizing (mean ± SE)"
  ) +
  theme(
    legend.position = "top"
  )
```

Only the measure changed, plus the `sex` level order and a `labs()` layer for
proper axis titles. The whole structure — filter, group, summarise, compute SE,
plot with dodged error bars — is reusable for any continuous measure in the
dataset.

Reusable enough that copy-pasting it starts to feel wrong. The
[functions deep-dive](module2-functions.qmd) turns it into a single line.
:::

## Take-home exercises

Six exercises to work through on your own time, using parts of the dataset the
session did not touch. They are loosely ordered by difficulty — ★ uses one or
two ideas from today, ★★★ combines most of them. Each has a worked solution;
try it properly before you open it.

All of them start from `data` and `levels_dict` as loaded in sections 1 and 3,
and the [dataset page](data.qmd) is your reference for what each column means.

### T1 · When did remote visits happen? ★

::: {.callout-note}
## Take-home 1
`ab_g_dyn__visit_type` records whether a visit was on-site, remote or hybrid —
but as the codes `1`, `2`, `3`. Label it using `levels_dict`, then count how many
visits of each type happened at each session. At which session were remote
visits most common?
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: th-1
visit_type_levels <- levels_dict |>
  filter(name == "ab_g_dyn__visit_type") |>
  arrange(order_level)

visits <- data |>
  mutate(
    visit_type = factor(
      ab_g_dyn__visit_type,
      levels = visit_type_levels$value,
      labels = visit_type_levels$label
    )
  )

visits |>
  count(session_id, visit_type) |>
  tidyr::pivot_wider(
    names_from = visit_type,
    values_from = n,
    values_fill = 0
  )
```

Every baseline visit was on-site. Remote visits peak sharply at `ses-03A`, the
3-year follow-up — the kind of pattern you see when outside events (a pandemic,
say) interrupt data collection partway through a study.

`pivot_wider()` spreads the counts into one column per visit type so the table
reads like a crosstab. `values_fill = 0` writes `0` rather than `NA` where a
combination never happened.
:::

### T2 · What does a remote visit cost? ★

::: {.callout-note}
## Take-home 2
Using your labelled `visit_type` from T1, work out the **percentage** of height,
weight and waist measurements that are missing for each visit type. Show it as a
bar chart.

Hint: `mean(is.na(x))` is the *proportion* missing — `is.na()` gives
`TRUE`/`FALSE`, and the mean of those is the share that are `TRUE`.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: th-2
#| fig-alt: "Bar chart of percentage missing for height, waist and weight by visit type; remote visits are around 80 percent missing for all three measures, while on-site and hybrid visits have no missing height or weight and about 6 percent missing waist."
#| fig-width: 10
#| fig-height: 5
visits |>
  group_by(visit_type) |>
  summarise(
    across(
      c(height = ph_y_anthr__height_mean,
        weight = ph_y_anthr__weight_mean,
        waist  = ph_y_anthr__waist_001),
      ~ 100 * mean(is.na(.x))
    )
  ) |>
  tidyr::pivot_longer(
    cols = height:waist,
    names_to = "measure",
    values_to = "pct_missing"
  ) |>
  ggplot(aes(x = visit_type, y = pct_missing, fill = measure)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_viridis_d() +
  labs(x = NULL, y = "% missing", fill = NULL) +
  theme_bw(20) +
  theme(legend.position = "top")
```

Remote visits are missing about 80% of their body measurements. On-site and
hybrid visits have no missing height or weight at all, and only about 6% missing
waist. Nobody can measure a
child's height over video.

This is **structured missingness**: whether a value is missing depends on
something else you know about the visit. Drop the incomplete rows and you drop
remote visits disproportionately — which, from T1, means you drop the 3-year
follow-up disproportionately. [Module 3](module3.qmd) takes this up.
:::

### T3 · Do parents and children agree about screen time? ★★

::: {.callout-note}
## Take-home 3
The dataset has four screen-time columns: parent- and youth-reported, weekday
and weekend. Make one figure showing the mean ± SE at each session, with
**reporter** as colour and **weekday/weekend** as facets.

Steps:

1. Select the ID columns and the four `_screen_` columns, and pivot them long.
2. From the old column name, make two new columns: `reporter` (`Parent` if the
   name contains `_p_`, otherwise `Youth`) and `day` (`Weekday` if it contains
   `wkdy`, otherwise `Weekend`). `stringr::str_detect()` tells you whether a
   string contains a pattern.
3. Drop values over 24 — there are only 24 hours in a day, and section 3 showed
   some are impossible.
4. Summarise and plot as in Q2.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: th-3
#| fig-alt: "Two panels, weekday and weekend, each showing mean screen time rising across five sessions for parent reports and youth reports; youth report consistently more than parents, and the gap widens over time."
#| fig-width: 12
#| fig-height: 6
data |>
  select(participant_id, session_id, matches("_screen_")) |>
  tidyr::pivot_longer(
    cols = matches("_screen_"),
    names_to = "what",
    values_to = "hours"
  ) |>
  mutate(
    reporter = if_else(stringr::str_detect(what, "_p_"), "Parent", "Youth"),
    day      = if_else(stringr::str_detect(what, "wkdy"), "Weekday", "Weekend"),
    visit    = stringr::str_replace(session_id, "ses-", "") |>
      readr::parse_number() |>
      factor()
  ) |>
  filter(hours <= 24) |>
  group_by(visit, reporter, day) |>
  summarise(
    n    = n(),
    mean = mean(hours),
    se   = sd(hours) / sqrt(n),
    .groups = "drop"
  ) |>
  ggplot(aes(x = visit, y = mean, colour = reporter, group = reporter)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    position = position_dodge(width = 0.4),
    width = .3
  ) +
  geom_line(position = position_dodge(width = 0.4)) +
  geom_point(position = position_dodge(width = 0.4), size = 4) +
  facet_grid(~ day) +
  scale_colour_manual(values = c(Parent = "grey40", Youth = "orange")) +
  labs(x = "Visit (years since baseline)", y = "Screen time, hours (mean ± SE)", colour = NULL) +
  theme_bw(20) +
  theme(legend.position = "top")
```

`filter(hours <= 24)` also removes `NA`s, because `NA <= 24` is `NA` and
`filter()` only keeps rows where the condition is `TRUE`. Handy, but worth knowing
that it happens.

Children report more screen time than their parents do at every visit, and the
gap grows as they get older — plausibly because parents see less of it. Which
report is "right" is not a question the data can answer, and it is exactly why
ABCD collects both.

The facets share a y-axis here (no `scales = "free"`), because both panels are in
hours and the weekday–weekend comparison is part of the point.
:::

### T4 · Household income and vocabulary at baseline ★★

::: {.callout-note}
## Take-home 4
At baseline only, compare the NIH Toolbox crystallized composite
(`nc_y_nihtb__comp__cryst__agecorr_score`) across the three household income
bands (`ab_g_dyn__cohort_income__hhold__3lvl`). Make a boxplot and a table of n,
mean and SD per band.

The income variable includes the codes `777` (decline to answer) and `999`
(don't know). Decide what to do with them, and write one sentence saying why.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: th-4
#| fig-alt: "Boxplots of NIH Toolbox crystallized score at baseline for three income bands, under 50k, 50k to 100k and over 100k, with medians rising across bands."
#| fig-width: 9
#| fig-height: 5.5
#| warning: false
income_levels <- levels_dict |>
  filter(name == "ab_g_dyn__cohort_income__hhold__3lvl") |>
  arrange(order_level)

baseline_income <- data |>
  filter(session_id == "ses-00A") |>
  mutate(
    income = factor(
      ab_g_dyn__cohort_income__hhold__3lvl,
      levels = income_levels$value,
      labels = income_levels$label
    ),
    cryst = nc_y_nihtb__comp__cryst__agecorr_score
  ) |>
  filter(income %in% c("< 50k", "50k to 100k", "> 100k")) |>
  mutate(income = droplevels(income))

baseline_income |>
  group_by(income) |>
  summarise(
    n    = sum(!is.na(cryst)),
    mean = round(mean(cryst, na.rm = TRUE), 1),
    sd   = round(sd(cryst, na.rm = TRUE), 1)
  )

baseline_income |>
  ggplot(aes(x = income, y = cryst)) +
  geom_boxplot(fill = "orange", col = "black") +
  labs(x = "Household income", y = "NIH Toolbox crystallized\n(age-corrected)") +
  theme_bw(20)
```

This solution **drops** `777` and `999`: they are not income bands, and they
cannot sit on an ordered low-to-high axis. That is a defensible choice, not the
only one — keeping them as their own category would let you check whether people
who decline to answer differ from everyone else, which matters if you later want
to generalise. Whatever you choose, say so in your methods.

`droplevels()` removes the now-empty factor levels, so the plot does not show
blank slots for them.

The medians rise with income. Whether that difference is bigger than chance
would produce, and whether it survives accounting for site, is a job for
[Module 3](module3.qmd) — a boxplot describes, it does not test.
:::

### T5 · Individual trajectories ★★

::: {.callout-note}
## Take-home 5
Mean ± SE plots hide the individuals. Plot mean cortical thickness
(`mr_y_smri__thk__dsk_mean`) against age for **30 randomly chosen
participants**, one faint line per person, with the average at each session
drawn on top.

Hints: pick the 30 IDs first with `distinct()` and `slice_sample()`, then
`filter(participant_id %in% ...)`. For one line per person, map
`group = participant_id`. A second layer can take its own data:
`geom_line(data = other_table, ...)`.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: th-5
#| fig-alt: "Thirty faint grey lines, one per participant, showing cortical thickness declining with age between about 9 and 16 years, with a thick orange line of session means on top that also slopes downward."
#| fig-width: 10
#| fig-height: 6
#| warning: false
set.seed(2026)   # so the random sample is the same every time

some_ids <- data |>
  distinct(participant_id) |>
  slice_sample(n = 30) |>
  pull(participant_id)

thickness <- data |>
  select(
    participant_id,
    session_id,
    age = ab_g_dyn__visit_age,
    thickness = mr_y_smri__thk__dsk_mean
  ) |>
  filter(!is.na(thickness))

session_means <- thickness |>
  group_by(session_id) |>
  summarise(
    age       = mean(age),
    thickness = mean(thickness)
  )

thickness |>
  filter(participant_id %in% some_ids) |>
  ggplot(aes(x = age, y = thickness)) +
  geom_line(aes(group = participant_id), colour = "grey60", alpha = 0.6) +
  geom_point(colour = "grey60", size = 1) +
  geom_line(data = session_means, colour = "orange", linewidth = 2) +
  geom_point(data = session_means, colour = "orange", size = 5) +
  labs(x = "Age (years)", y = "Mean cortical thickness (mm)") +
  theme_bw(20)
```

`set.seed()` fixes R's random number generator, so `slice_sample()` picks the
same 30 people every run — essential for anything you want to reproduce.

MRI was only collected at baseline, year 2 and year 4, which is why each person
has at most three points (and the average has three). Nearly every individual
line slopes downward, at broadly similar rates, but they start at very
different heights: people differ a lot in *level*, rather less in *slope*. That observation is the core idea
behind the mixed-effects models used for longitudinal data — a reason Module 3
cannot treat these rows as independent.
:::

### T6 · A "Table 1" ★★★

::: {.callout-note}
## Take-home 6
Almost every paper opens with a table describing the sample. Build one for the
baseline visit, **by sex**: n, and mean (SD) of age, height, weight, youth
weekday screen time, and CBCL internalizing and externalizing.

Aim for one row per variable and one column per sex, with cells that read like
`10.1 (0.6)`.

Hint: this is `summarise(across(..., list(mean = ..., sd = ...)))`, then a
`pivot_longer()` and a `pivot_wider()`. `names_sep` in `pivot_longer()` can split
a name like `age_mean` into two columns.
:::

::: {.callout-tip collapse="true"}
## Solution
```{r}
#| label: th-6
baseline <- data |>
  filter(session_id == "ses-00A") |>
  transmute(
    sex = case_when(
      ab_g_stc__cohort_sex == 1 ~ "Male",
      ab_g_stc__cohort_sex == 2 ~ "Female"
    ),
    age      = ab_g_dyn__visit_age,
    height   = ph_y_anthr__height_mean,
    weight   = ph_y_anthr__weight_mean,
    screen   = nt_y_stq__screen__wkdy_sum,
    cbclint  = mh_p_cbcl__synd__int_sum,
    cbclext  = mh_p_cbcl__synd__ext_sum
  )

counts <- baseline |> count(sex)
counts

baseline |>
  group_by(sex) |>
  summarise(
    across(
      age:cbclext,
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        sd   = ~ sd(.x, na.rm = TRUE)
      )
    )
  ) |>
  tidyr::pivot_longer(
    cols = -sex,
    names_to = c("variable", "stat"),
    names_sep = "_"
  ) |>
  tidyr::pivot_wider(names_from = stat, values_from = value) |>
  mutate(cell = sprintf("%.1f (%.1f)", mean, sd)) |>
  select(variable, sex, cell) |>
  tidyr::pivot_wider(names_from = sex, values_from = cell)
```

Three reshapes in a row, so step through it: run the pipeline one line at a time
and `glimpse()` after each step.

- `across()` with a named `list()` makes columns like `age_mean`, `age_sd`.
- `pivot_longer(names_sep = "_")` splits each name at the `_` into `variable` and
  `stat`. That is why the new columns are named `cbclint` rather than
  `cbcl_int`: an extra underscore would split in the wrong place.
- The first `pivot_wider()` puts `mean` and `sd` side by side so `sprintf()` can
  format them into `"10.1 (0.6)"`. `%.1f` means "a number with one decimal
  place", and unlike `round()` it keeps a trailing zero: `5.0`, not `5`.
- The second `pivot_wider()` gives one column per sex.

`transmute()` is `mutate()` that keeps only the columns it creates — a
compact select-and-rename for when every column needs a new name.

For a table going straight into a manuscript, packages such as **gtsummary**
build this in one call. Building it by hand once is how you learn what those
packages are doing.
:::

## What you can now do

- Read an `.RDS` file and explain why it beats a CSV for R-to-R work
- Call a function with `package::function()` without loading the package
- Pick columns by position, name and pattern, renaming as you go
- Recode values with `case_when()`, and transform many columns at once with `across()`
- Attach labels from a lookup table with `left_join()`, and check that the join worked
- Scan for implausible values before trusting a variable
- Reshape wide data to long with `pivot_longer()`, and facet by the result
- Control fill, colour, factor order, palettes, themes and legend position
- Compute n, mean, SD and SE by group, and plot means with error bars

## Recording

The recording of the live session on Sep 24, 2026 will be posted here after the
workshop.

## Going further

Optional, and none of it is assumed in Module 3:

- [Wrangling deep-dive](module2-wrangling.qmd) — select helpers, `across()`,
  recoding, joins, reshaping, and labelling every coded variable at once
- [ggplot2 anatomy deep-dive](module2-ggplot-anatomy.qmd) — fill vs colour,
  palettes, facets, dodging, themes, and saving figures
- [Writing functions deep-dive](module2-functions.qmd) — turn today's pipelines
  and plots into reusable functions
- The [plot gallery](module1-plot-gallery.qmd) — more geoms: density, violin,
  scatter + trend line, spaghetti plots
- The [dataset reference](data.qmd) — every variable, every coded level, and why
  the data is missing where it is
- A [past workshop on this material](https://www.youtube.com/watch?v=7t9JhBJm1pg)
- [R for Data Science](https://r4ds.hadley.nz), the chapters on data
  visualization, data tidying and joins
- The [ggplot2 cheatsheet](https://rstudio.github.io/cheatsheets/data-visualization.pdf)

Next: [Module 3 · Inferential Statistics →](module3.qmd) on Oct 29, where the
question moves from "what does the data look like?" to "is this difference
real?".
