---
title: "Deep-dive · Wrangling ABCD-Shaped Data"
subtitle: "Optional companion to Module 2"
pagetitle: "Wrangling: Joins, across() & Pivots | ABCD R Workshop"
description-meta: "Select helpers, across(), case_when(), left_join() and pivot_longer() in R, applied to ABCD-shaped data — including labelling every coded variable at once."
open-graph:
  description: "Select helpers, across(), recoding, joins and pivots in dplyr and tidyr, on ABCD-shaped data."
twitter-card:
  description: "Select helpers, across(), recoding, joins and pivots in dplyr and tidyr, on ABCD-shaped data."
editor_options:
  chunk_output_type: console
---

This page expands on the data-wrangling steps in [Module 2](module2.qmd). Nothing
here is needed for the session; it is for when you want to know what else the
tools can do, or why they behave the way they do.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(readr)

data <- readRDS("data/data.RDS")
levels_dict <- read_csv("data/data_dictionary_levels.csv")
```

## Choosing columns

Module 2 used `select()` by position, by name and by `matches()`. There are
several other **selection helpers**, and they all work anywhere dplyr expects a
set of columns — `select()`, `across()`, `pivot_longer(cols = ...)`:

| Helper | Keeps columns whose name… |
|:--|:--|
| `starts_with("nc_")` | begins with `nc_` |
| `ends_with("_sum")` | ends with `_sum` |
| `contains("cbcl")` | contains `cbcl` anywhere, taken literally |
| `matches("_(wkdy\|wknd)_")` | matches a *regular expression* |
| `where(is.numeric)` | — this one tests the column's **contents**, not its name |

ABCD's naming scheme makes these very effective. Every NIH Toolbox score:

```{r}
#| label: starts-with
data |>
  select(starts_with("nc_")) |>
  names()
```

Every parent-reported variable — the second chunk of the name is the reporter:

```{r}
#| label: parent-reported
data |>
  select(matches("^[a-z]{2}_p_")) |>
  names()
```

`^[a-z]{2}_p_` reads as "at the start, two lowercase letters, then `_p_`". You do
not need to become fluent in regular expressions, but the handful that match
ABCD's structure are worth keeping in a note.

Helpers combine. `-` removes, and `&` means both:

```{r}
#| label: combine-helpers
data |>
  select(starts_with("mh_") & ends_with("_sum"), -contains("bisbas")) |>
  names()
```

## Doing the same thing to many columns: `across()`

`across()` applies one function to several columns inside `mutate()` or
`summarise()`. It takes two arguments: *which* columns, and *what to do*.

Module 2 wrote the function as `~ round(.x)`. That `~` form is a compact way to
write a small function on the spot, with `.x` standing for each column in turn.
When the function needs no extra arguments you can just name it:

```{r}
#| label: across-forms
data |>
  summarise(across(starts_with("nt_"), mean)) |>
  glimpse()

data |>
  summarise(across(starts_with("nt_"), ~ mean(.x, na.rm = TRUE))) |>
  glimpse()
```

The first gives `NA`s — `mean()` meets missing values and, as Module 1 showed,
refuses to guess. The second passes `na.rm = TRUE`, and extra arguments are
exactly the case the `~ .x` form exists for.

Several summaries at once, with a named list:

```{r}
#| label: across-list
data |>
  group_by(session_id) |>
  summarise(
    across(
      c(nt_y_stq__screen__wkdy_sum, nt_y_stq__screen__wknd_sum),
      list(mean = ~ mean(.x, na.rm = TRUE), sd = ~ sd(.x, na.rm = TRUE))
    )
  )
```

The new columns are named `{column}_{function}` automatically.

## Recoding values

### `case_when()`

Module 2 recoded `sex` with `case_when()`. Each line is `condition ~ result`.
Lines are tried **top to bottom**, and the first match wins — so order matters
when conditions overlap:

```{r}
#| label: case-when
data |>
  mutate(
    screen_band = case_when(
      nt_y_stq__screen__wkdy_sum > 24 ~ "implausible",
      nt_y_stq__screen__wkdy_sum >= 6 ~ "6+ h",
      nt_y_stq__screen__wkdy_sum >= 3 ~ "3–6 h",
      nt_y_stq__screen__wkdy_sum >= 0 ~ "under 3 h",
      .default = NA
    )
  ) |>
  count(screen_band)
```

Put the `> 24` line last and it would never fire: those rows would already have
matched `>= 6`.

### `factor(levels, labels)`

For coded variables with a dictionary, the [dataset page](data.qmd#labelling-one-variable)
shows a different route: hand the codes and labels to `factor()`. This has two
advantages over `case_when()` — you do not have to type the labels, so there is
nothing to mistype, and the result is a factor that keeps the dictionary's
order.

`case_when()` is the right tool when you are *making* categories (like the
screen-time bands above). `factor()` is the right tool when a dictionary
already defines them.

## Labelling every coded variable at once

The dataset page labels one variable at a time and notes that doing all
thirteen in one stroke needs more tools. Here they are.

First, write the one-variable recipe as a **function** — a reusable piece of
code with a name:

```{r}
#| label: label-function
apply_labels <- function(x, var) {
  lv <- levels_dict |>
    filter(name == var) |>
    arrange(order_level)

  factor(x, levels = lv$value, labels = lv$label)
}
```

`function(x, var)` declares two inputs: `x`, the column's values, and `var`, the
column's name, used to look up its levels. The last line inside the braces is
what the function returns.

Then apply it with `across()`. `any_of()` selects every column that appears in
the dictionary, and `cur_column()` tells the function which column it is
currently working on:

```{r}
#| label: label-all
data_labelled <- data |>
  mutate(
    across(
      any_of(unique(levels_dict$name)),
      ~ apply_labels(.x, cur_column())
    )
  )

data_labelled |> count(ab_g_dyn__visit_type)
data_labelled |> count(ab_g_dyn__cohort_income__hhold__3lvl)
```

Thirteen variables, one pipeline. Check the labelling did not invent missing
values — a code missing from the dictionary would silently become `NA`:

```{r}
#| label: label-check
coded <- intersect(names(data), levels_dict$name)

tibble(
  variable = coded,
  na_before = sapply(coded, \(v) sum(is.na(data[[v]]))),
  na_after  = sapply(coded, \(v) sum(is.na(data_labelled[[v]])))
)
```

Identical counts: every code found its label. (`\(v)` is shorthand for
`function(v)`.)

## Joins

A **join** combines two tables by matching values in a key column. Module 2 used
`left_join()` to attach site names. The family:

| Join | Keeps |
|:--|:--|
| `left_join(x, y)` | every row of `x`; columns from `y` where they match, `NA` where they do not |
| `inner_join(x, y)` | only rows that match in both |
| `full_join(x, y)` | every row of both |
| `anti_join(x, y)` | rows of `x` with **no** match in `y` — the one for finding problems |

`left_join()` is the everyday choice for "look up extra information". It never
loses a row of your data, so a failed match shows up as an `NA`, not as a
participant silently disappearing.

`anti_join()` is the checking tool. Which site codes in the data have no entry
in the dictionary?

```{r}
#| label: anti-join
site_levels <- levels_dict |>
  filter(name == "ab_g_dyn__design_site") |>
  transmute(value = as.character(value), site_label = label)

data |>
  mutate(site = as.character(ab_g_dyn__design_site)) |>
  distinct(site) |>
  anti_join(site_levels, join_by(site == value))
```

Zero rows: every site matched.

### Why Module 2 converted both keys to character

Keys must be the same type. In the data, site is a **factor**; in the levels
dictionary, `value` is a **number**. Try joining them as they are:

```{r}
#| label: join-type-error
#| error: true
data |>
  left_join(
    levels_dict |> filter(name == "ab_g_dyn__design_site"),
    join_by(ab_g_dyn__design_site == value)
  )
```

The error is helpful: it names both columns and both types. Converting each side
with `as.character()` gives them a common type that preserves the codes exactly.

::: {.callout-warning}
## Do not use `as.numeric()` on a factor
It is tempting to convert the factor to a number instead. But `as.numeric()` on a
factor returns the level's **position**, not the value printed on it. Here the
levels happen to be `"1"` to `"22"` in order, so it would work by luck. With
levels like `"777"` or out-of-order codes, it quietly gives wrong numbers. Go via
`as.character()` first, or join as text.
:::

## Reshaping: long and wide

`pivot_longer()` stacks columns into rows; `pivot_wider()` spreads them back. The
two are inverses.

```{r}
#| label: pivot-longer
anthro_long <- data |>
  select(participant_id, session_id, starts_with("ph_y_anthr")) |>
  pivot_longer(
    cols = starts_with("ph_y_anthr"),
    names_to = "measure",
    values_to = "value"
  )

anthro_long
```

Three times the rows, three columns fewer. Long data is what ggplot2 wants for
faceting and colouring by "which measure", and what `group_by()` wants for "the
same summary for each measure":

```{r}
#| label: long-summary
anthro_long |>
  group_by(measure) |>
  summarise(
    n_present = sum(!is.na(value)),
    mean      = mean(value, na.rm = TRUE)
  )
```

`pivot_wider()` goes the other way. A common longitudinal use: one row per
participant, one column per visit — handy for looking at each person's
trajectory, and the shape a paired test in Module 3 will want:

```{r}
#| label: pivot-wider
data |>
  select(participant_id, session_id, nt_y_stq__screen__wkdy_sum) |>
  pivot_wider(
    names_from = session_id,
    values_from = nt_y_stq__screen__wkdy_sum
  )
```

The `NA`s on the right are participants who had left the study by that visit —
the attrition described on the [dataset page](data.qmd#what-is-missing-and-why),
now visible as gaps in a row rather than as absent rows.

Back to [Module 2](module2.qmd).
