---
title: "Deep-dive · Anatomy of a ggplot2 Figure"
subtitle: "Optional companion to Module 2"
pagetitle: "ggplot2 Anatomy: Scales, Facets & Themes | ABCD R Workshop"
description-meta: "How a ggplot2 figure is built: fill vs colour, manual and viridis palettes, facet_grid vs facet_wrap, dodging, themes, labels, and saving with ggsave()."
open-graph:
  description: "Fill vs colour, palettes, facets, dodging, themes and ggsave() — the parts of a publication-ready ggplot2 figure."
twitter-card:
  description: "Fill vs colour, palettes, facets, dodging, themes and ggsave() — the parts of a publication-ready ggplot2 figure."
editor_options:
  chunk_output_type: console
---

[Module 2](module2.qmd) built its figures quickly, in a live session. This page
slows down and takes the pieces apart. For a tour of *which* geom to use for
which kind of data, see the [plot gallery](module1-plot-gallery.qmd); this page
is about the parts every figure shares.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)

data <- readRDS("data/data.RDS")

anthro <- data |>
  mutate(
    sex = case_when(
      ab_g_stc__cohort_sex == 1 ~ "Male",
      ab_g_stc__cohort_sex == 2 ~ "Female"
    ) |> factor(levels = c("Male", "Female")),
    visit = stringr::str_replace(session_id, "ses-", "") |>
      readr::parse_number(),
    visit = paste0("yr-", visit)
  ) |>
  select(
    participant_id, visit, sex,
    age    = ab_g_dyn__visit_age,
    height = ph_y_anthr__height_mean,
    weight = ph_y_anthr__weight_mean,
    waist  = ph_y_anthr__waist_001
  )
```

## A figure is layers

Every ggplot2 figure is the same few components, added with `+`:

| Component | Job | Module 2 example |
|:--|:--|:--|
| data + `aes()` | which columns map to which visual property | `aes(x = visit, y = waist, fill = sex)` |
| `geom_*()` | what marks to draw | `geom_boxplot()` |
| `scale_*()` | how values become colours, positions, sizes | `scale_fill_manual()` |
| `facet_*()` | split into panels | `facet_grid(~ what)` |
| `coord_*()` | the coordinate system | `coord_flip()` |
| `theme*()`, `labs()` | everything that is not data: fonts, grid, legend, titles | `theme_bw(20)` |

You can add them in almost any order. Module 2 put `theme_bw(20)` before the
geom in one plot and after it in another, and the results are the same. The one
exception: `theme()` tweaks must come **after** a complete theme like
`theme_bw()`, or the complete theme overwrites them.

## Fill vs colour

Two aesthetics, easily confused:

- **`colour`** (or `col`, or `color`) — lines, points, and the *outline* of shapes
- **`fill`** — the *inside* of shapes: bars, boxes, violins, ribbons, and points
  with shapes 21–25

```{r}
#| label: fig-fill-colour
#| fig-cap: "Left: sex mapped to fill. Right: sex mapped to colour."
#| fig-alt: "Two boxplots of height by sex. The left has orange and grey filled boxes with black outlines; the right has white boxes with orange and grey outlines."
#| layout-ncol: 2
#| warning: false
anthro |>
  ggplot(aes(x = sex, y = height, fill = sex)) +
  geom_boxplot(col = "black") +
  scale_fill_manual(values = c("orange", "grey60"))

anthro |>
  ggplot(aes(x = sex, y = height, colour = sex)) +
  geom_boxplot() +
  scale_colour_manual(values = c("orange", "grey60"))
```

That is why Module 2 mapped **both** `fill = sex` and `col = sex`, then set
`col = "black"` in `geom_boxplot()`: fill carries the grouping, and a fixed black
outline keeps the boxes crisp. Setting an aesthetic *inside* `aes()` maps it to a
column; setting it *outside* fixes it to one value for the whole layer.

## Scales: choosing the colours

Every mapped aesthetic has a scale, and every scale has a default. You override
it with a `scale_<aesthetic>_<type>()`:

- **`scale_fill_manual(values = c(...))`** — you choose. Colours are handed out
  in the order of the factor's levels, so fix the level order first (Module 2's
  "why are the colours swapped?" callout is this rule in action). Name them to
  make the mapping explicit and immune to ordering:
  `values = c(Male = "orange", Female = "grey60")`.
- **`scale_fill_viridis_d()`** — the viridis palettes: perceptually even, legible
  in greyscale, and distinguishable with the common forms of colour blindness.
  `_d` for discrete data, `_c` for continuous.

```{r}
#| label: fig-viridis
#| fig-cap: "Five visits coloured with the viridis palette"
#| fig-alt: "Overlapping density curves of weight at each of five visits, shaded from dark purple at year 0 to yellow at year 4."
#| warning: false
anthro |>
  ggplot(aes(x = weight, fill = visit)) +
  geom_density(alpha = 0.6, colour = NA) +
  scale_fill_viridis_d() +
  theme_bw(14)
```

With five ordered categories, a sequential palette like viridis reads as "earlier
to later" at a glance, which a set of arbitrary colours would not.

## Facets

`facet_grid()` and `facet_wrap()` both split a plot into panels by a column.

- **`facet_grid(rows ~ cols)`** lays panels on a strict grid. `~ what` means
  "columns by `what`, no rows"; `sex ~ what` would give a grid of sex × measure.
- **`facet_wrap(~ what)`** lays panels in a line that wraps, like text. Better
  when there are many panels, and you can set `ncol` or `nrow`.

`scales =` controls whether panels share axes: `"fixed"` (the default),
`"free_x"`, `"free_y"`, or `"free"` for both.

```{r}
#| label: fig-facet-wrap
#| fig-cap: "Growth curves, one panel per measure, with free y-axes"
#| fig-alt: "Three scatterplots of height, waist and weight against age, each with its own y-axis range and a smooth trend line per sex."
#| fig-width: 11
#| fig-height: 4
#| warning: false
anthro |>
  pivot_longer(height:waist, names_to = "measure", values_to = "value") |>
  ggplot(aes(x = age, y = value, colour = sex)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(se = FALSE, method = "loess", formula = y ~ x) +
  facet_wrap(~ measure, scales = "free_y") +
  scale_colour_manual(values = c(Male = "orange", Female = "grey40")) +
  theme_bw(14)
```

Free scales are right when the panels are in different units, as here. When the
panels share a unit and you want readers to compare *across* them, keep the
scales fixed. Otherwise identical-looking panels can hide very different values.

::: {.callout-note}
## `coord_flip()` and free scales
In Module 2, `facet_grid(~ what, scales = "free")` combined with `coord_flip()`
gave each panel its own *value* axis, even though the value axis ended up
horizontal. Scales belong to the aesthetic (`y = value`), not to the screen
direction, so they flip along with the coordinates.
:::

## Dodging

When two groups share an x position, their marks land on top of each other.
**Position adjustments** move them:

- `position_dodge(width = )` — side by side. Module 2 used it for points, lines
  and error bars.
- `position_jitter()` — a small random shift, for showing raw points without
  overplotting (`geom_jitter()` is shorthand).
- `position_stack()` / `position_fill()` — stacked bars, the default for
  `geom_bar()`.

For dodging, the `width` must match across every layer that should line up.
Build it once and reuse it:

```{r}
#| label: fig-dodge
#| fig-cap: "One dodge object shared by three layers"
#| fig-alt: "Mean height by visit for males in orange and females in grey, points and error bars offset side by side, joined by lines."
#| warning: false
dodge <- position_dodge(width = 0.5)

anthro |>
  group_by(visit, sex) |>
  summarise(
    n    = sum(!is.na(height)),
    mean = mean(height, na.rm = TRUE),
    se   = sd(height, na.rm = TRUE) / sqrt(n),
    .groups = "drop"
  ) |>
  ggplot(aes(x = visit, y = mean, colour = sex, group = sex)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2, position = dodge) +
  geom_line(position = dodge) +
  geom_point(size = 3, position = dodge) +
  scale_colour_manual(values = c(Male = "orange", Female = "grey40")) +
  theme_bw(14)
```

## Themes and labels

`theme_bw(base_size = 20)` sets a complete look and a base font size in one call;
every text element scales from it. Other complete themes: `theme_minimal()`,
`theme_classic()` (no grid, a common journal style), `theme_light()`.

`theme()` then adjusts individual elements. The ones you will reach for most:

```r
theme(
  legend.position = "top",             # or "bottom", "none", or c(0.9, 0.1)
  panel.grid.minor = element_blank(),  # remove minor grid lines
  strip.background = element_blank(),  # plain facet labels
  axis.text.x = element_text(angle = 45, hjust = 1)
)
```

`labs()` sets every piece of text that describes the data. A figure headed for a
paper should never ship with a raw variable name on an axis:

```{r}
#| label: fig-labelled
#| fig-cap: "A finished figure"
#| fig-alt: "Horizontal boxplots of waist circumference by visit and sex with a title, subtitle, axis labels in inches, and a caption noting the data is synthetic."
#| fig-width: 9
#| fig-height: 5.5
#| warning: false
anthro |>
  ggplot(aes(x = visit, y = waist, fill = sex)) +
  geom_boxplot(col = "black", outlier.size = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c(Male = "orange", Female = "grey60")) +
  labs(
    x        = NULL,
    y        = "Waist circumference (inches)",
    fill     = NULL,
    title    = "Waist circumference increases across visits",
    subtitle = "Boxes show median and interquartile range",
    caption  = "Synthetic ABCD-shaped data, n = 500"
  ) +
  theme_bw(14) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

`x = NULL` removes an axis title entirely, which is often cleaner than an empty
string. `fill = NULL` does the same for the legend title.

## Saving a figure

`ggsave()` writes the most recent plot to disk. The file extension picks the
format:

```r
ggsave("output/waist-by-visit.png", width = 7, height = 4.5, dpi = 300)
ggsave("output/waist-by-visit.pdf", width = 7, height = 4.5)
```

Two things make the difference between a figure that looks right on screen and
one that looks right on the page:

- **Set `width` and `height` explicitly**, in inches, at the size the figure will
  actually be printed. The text sizes from `theme_bw(base_size)` are in points,
  so a plot saved at poster size and shrunk into a paper column ends up with
  unreadable text.
- **Use `dpi = 300`** for raster formats (PNG, TIFF) — the usual minimum for
  print. Vector formats (PDF, SVG) stay sharp at any size and ignore `dpi`.

To save a specific plot rather than the last one, store it and pass it in:
`p <- ggplot(...) + ...`, then `ggsave("output/p.png", plot = p, ...)`.

Back to [Module 2](module2.qmd).
